---
title: "Assessing the impact of COVID-19 pandemic on hip replacement activity."
author: "Alexander Lawless"
date: "21nd July 2021"
output: 
  html_document:
    theme: cerulean
    toc: TRUE
    toc_float: TRUE
    toc_collapsed: TRUE
    toc_depth: 4
    number_sections: TRUE
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9, fig.height = 6)

library(tidyverse)
library(odbc)
library(DBI)
library(plotly)
library(lubridate)
library(ggrepel)
library(dygraphs)
library(xts) 
library(janitor)
library(knitr)
library(PropCIs)
library(patchwork)
library(kableExtra)
library(DT)
library(scales)

library(streamgraph)

inline_hook <- function(x) {
 
  if(is.numeric(x)) {
    formatted <- format(x, digits = 2)
    
  } else {
    formatted <- x
  }
  
   paste0("**", formatted, "**")
  
}

knit_hooks$set(inline = inline_hook)

create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                               lengthMenu = list(c(7, 10,50,-1),
                                                 c(7, 10,50,"All"))))
}

## Set SU theme ####
My_rgb2hex <- function(r,g,b) sprintf('#%s',paste(as.hexmode(c(r,g,b)),collapse = ''))

SU_colours <- c (
  `orange` = My_rgb2hex(248,191,7),# "#f9bf07",
  `charcoal` = My_rgb2hex(44,40,37),# "#2c2825",
  `slate` = My_rgb2hex(104,111,115), # "#686f73",
  `blue` = My_rgb2hex(88,29,193), # "#5881c1",
  `red` = My_rgb2hex(236,101,85), # "#ec6555",
  #additional accent colours from word doc template
  `yellow` = My_rgb2hex(252,229,155),
  `grey` = My_rgb2hex(163,168,172),
  `white` = My_rgb2hex(255,255,255),
  #light and dark ends from colour theme in word doc
  `light orange` = My_rgb2hex(253,242,205),
  `dark orange` = My_rgb2hex(124,95,3),
  `light charcoal` = My_rgb2hex(235,233,231),
  `dark charcoal` = 	"#000000",#black
  `light slate` = My_rgb2hex(224,226,227),
  `dark slate` = My_rgb2hex(51,55,57),
  `light blue` = My_rgb2hex(221,229,242),
  `dark blue` = My_rgb2hex(38,61,102),
  `light red` = My_rgb2hex(251,224,220),
  `dark red` = My_rgb2hex(144,29,16),
  `light yellow` = My_rgb2hex(254,249,235),
  `dark yellow` = My_rgb2hex(197,152,5),
  `light grey`=My_rgb2hex(236,237,238),
  `dark grey` = My_rgb2hex(79,84,88),
  `light white`=My_rgb2hex(242,242,242),
  `dark white` =My_rgb2hex(127,127,127),
  `red2` = My_rgb2hex(215,25,28),
  `orange2` = My_rgb2hex(253,174,97),
  `yellow2` = My_rgb2hex(255,255,191),
  `green2` = My_rgb2hex(171,221,164),
  `blue2` = My_rgb2hex(43,131,186)
)



SU_cols <- function(...) {
  cols <- c(...)
  
  if (is.null(cols))
    return (SU_colours)
  
  SU_colours[cols]
}

SU_palettes <- list(
  `main` = SU_cols("orange","charcoal","slate","blue","red"),
  `oranges` = SU_cols("light orange","orange","dark orange"),
  `slates` = SU_cols("light slate","slate","dark slate"),
  `mixed` = SU_cols("dark red","orange","yellow","light blue","slate"),
  `oj_coal` = SU_cols("yellow","orange","red","dark red","dark charcoal"),
  `oj_red` = SU_cols("yellow","orange","red","dark red"),
  `white_oj_coal` = SU_cols("white","yellow","orange","red","dark red","dark charcoal"),#added since shared
  `lyellow_oj_coal` = SU_cols("light yellow","orange","red","dark red","dark charcoal"),#added since shared
  `wy_oj_coal` = SU_cols("white","light yellow","yellow","orange","red","dark red","charcoal","dark charcoal"),
  `red_coal` = SU_cols("red","dark red","charcoal","dark charcoal"),
  `blue_yellow_red` = SU_cols("red2","orange2","yellow2","green2","blue2"),
  `red_yellow_blue` = SU_cols("blue2","green2","yellow2","orange2","red2")
)

SU_pal <- function(palette = "main", reverse = FALSE, ...) {
  pal <- SU_palettes[[palette]]
  
  if (reverse) pal <- rev(pal)
  
  colorRampPalette(pal, ...)
}   

scale_color_SU <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
  pal <- SU_pal(palette = palette, reverse = reverse)
  
  if (discrete) {
    discrete_scale("colour", paste0("SU_", palette), palette = pal, ...)
  } else {
    scale_color_gradientn(colours = pal(256), ...)
  }
}


scale_fill_SU <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
  pal <- SU_pal(palette = palette, reverse = reverse)
  
  if (discrete) {
    discrete_scale("fill", paste0("SU_", palette), palette = pal, ...)
  } else {
    scale_fill_gradientn(colours = pal(256), ...)
  }
}  

##Theme


##set theme settings
theme_SU<-
  function (base_size){
    theme_minimal(
      #base_family = "Segoe UI", 
      base_size=12
      ) %+replace% 
      theme(axis.title = element_text(size=11, face="bold",colour=SU_cols("charcoal")),
            plot.title = element_text(hjust=0,face="bold",size=12,colour=SU_cols("charcoal"),margin=margin(b=4,unit="pt")),
            plot.subtitle = element_text(hjust=0,face="italic",size=10,colour=SU_cols("charcoal"),margin=margin(b=4,unit="pt")),
            plot.caption = element_text(hjust = 0,face="italic",size=9,colour=SU_cols("slate"),margin=margin(b=4,unit="pt")),
            legend.text = element_text(size=10,colour=SU_cols("charcoal")),
            legend.title = element_text(face="bold",size=11,colour=SU_cols("charcoal"),margin=margin(b=4,unit="pt")))
  }



##set theme by 
theme_set(theme_SU())



## A. Import SUS data ####

# Connection
a.con <- dbConnect(odbc(), 
                 Driver = "SQL Server", 
                 server = "PRODNHSESQL101", 
                 Database = "NHSE_Sandbox_Spec_Neurology")

a_lsoa_to_stp <- "https://opendata.arcgis.com/datasets/d30531b5888a4e34be4746399d696409_0.csv" %>%
  read_csv() %>% 
  clean_names()

a_stp_to_nhsr <- "https://opendata.arcgis.com/datasets/888dc5cc66ba4ad9b4d935871dcce251_0.csv" %>%
  read_csv() %>% 
  clean_names()


# Import SUS inpatient hip activity (2018 onwards)
a.ip_hip <-
  as_tibble(
    dbGetQuery(a.con, '
select *
from [NHSE_Sandbox_Spec_Neurology].[dbo].[HF_covid_hip_activity_2]')) %>% 
  mutate(procedure_type = case_when(
    (totalHipReplacementPrimary == 1 |
      totalProstheticReplacementHeadFemurPrimary == 1 |
      hybridProstheticHipReplacementPrimary == 1 |
      otherHipReplacementPrimary == 1 |
      resurfacingArthroplastyJointPrimary == 1 ) ~ "Primary",
    (totalHipReplacementRevision == 1 |
       totalProstheticReplacementHeadFemurRevision == 1 |
       hybridProstheticHipReplacementRevision == 1 |
       otherHipReplacementRevision == 1|
       resurfacingArthroplastyJointRevision == 1) ~ "Revision",
    TRUE ~ "NA" )) %>% 
  mutate(Age_range = case_when(Age_at_Start_of_Episode_SUS >= 0 & Age_at_Start_of_Episode_SUS < 20 ~ "0-19",
                               (Age_at_Start_of_Episode_SUS >= 20 & Age_at_Start_of_Episode_SUS <40) ~  "20-39",
                               (Age_at_Start_of_Episode_SUS >= 40 & Age_at_Start_of_Episode_SUS <60) ~  "40-59",
                               (Age_at_Start_of_Episode_SUS >= 60 & Age_at_Start_of_Episode_SUS <80) ~  "60-79",
                               (Age_at_Start_of_Episode_SUS >= 80 & Age_at_Start_of_Episode_SUS <100) ~ "80-99",
                               Age_at_Start_of_Episode_SUS >= 100 ~ "100+")) %>% 
  mutate(Age_range = factor(Age_range, levels = c("0-19", 
                                                  "20-39",
                                                  "40-59",
                                                  "60-79",
                                                  "80-99", 
                                                  "100+"))) %>% 
  left_join(a_lsoa_to_stp, by = c("Der_Postcode_LSOA_2011_Code" = "lsoa11cd")) %>% 
  left_join(a_stp_to_nhsr %>% 
              select(stp20cd, nhser20cd, nhser20nm) %>% 
              distinct(), 
            by = c("stp20cd"))

a.ip_hip_unplanned <-
  as_tibble(
    dbGetQuery(a.con, '
select *
from [NHSE_Sandbox_Spec_Neurology].[dbo].[HF_covid_hip_activity_ip_unplanned]'))

a.opa_hip <-
as_tibble(
  dbGetQuery(a.con, '
select *
from [NHSE_Sandbox_Spec_Neurology].[DBO].[HF_covid_hip_activity_OPA]'))

a.ed_hip <-
  as_tibble(
    dbGetQuery(a.con, '
select *
from [NHSE_Sandbox_Spec_Neurology].[DBO].[HF_covid_hip_activity_ED]'))


a.ed_diabetes <-
  as_tibble(
    dbGetQuery(a.con, '
select *
from [NHSE_Sandbox_Spec_Neurology].[DBO].[HF_covid_diabetes_activity_ED]'))

# Disconnect! # 
dbDisconnect(a.con) 

## B.1. Time series summaries ####
# Monthly 
# Inpatients
b_hip_ip_month <-
a.ip_hip %>% 
  group_by(substr(Admission_Date, 1, 7)) %>% #reduce discharge date to discharge month
  summarise(n_admissions = n_distinct(APCS_Ident)) %>%  #sum monthly discharges
  rename(Month = 1) %>% 
  mutate(Month = paste0(Month,"-01")) %>% #Add day value for as.Date function
  mutate(Month = as.Date(Month)) %>% #mutate from chr to date
  filter(Month > 2018-01) 
# Outpatients
b_hip_opa_month <-
a.opa_hip %>% 
  group_by(substr(Appointment_Date, 1, 7)) %>% #reduce discharge date to discharge month
  summarise(n_appointments = n_distinct(OPA_Ident)) %>%  #sum monthly appointments
  rename(Month = 1) %>% 
  mutate(Month = paste0(Month,"-01")) %>% #Add day value for as.Date function
  mutate(Month = as.Date(Month)) %>% #mutate from chr to date
  filter(Month > 2018-01) 
# ED
b_hip_ed_month <-
a.ed_hip %>% 
  group_by(substr(Arrival_Date, 1, 7)) %>% #reduce discharge date to discharge month
  summarise(n_attendances = n_distinct(EC_Ident)) %>%  #sum monthly attendances
  rename(Month = 1) %>% 
  mutate(Month = paste0(Month,"-01")) %>% #Add day value for as.Date function
  mutate(Month = as.Date(Month)) %>% #mutate from chr to date
  filter(Month > 2018-01)
# Merge
b_hip_all_month <-
  b_hip_ip_month %>% 
  left_join(b_hip_opa_month, by = ("Month")) %>% 
  left_join(b_hip_ed_month, by = ("Month")) %>% 
  pivot_longer(cols = -Month, names_to = "Activity", values_to = "Count") %>% 
  mutate(Activity = case_when(Activity == "n_admissions" ~ "Inpatient",
                              Activity == "n_appointments" ~ "Outpatient",
                              Activity == "n_attendances" ~ "Emergency Care",
                              TRUE ~ "NA")) %>% 
  mutate(Activity = factor(Activity, levels = c("Inpatient", "Outpatient", "Emergency Care")))

# Daily
# Inpatient
b_hip_ip_daily <-
  a.ip_hip %>% 
  group_by(Admission_Date) %>% 
  summarise(n_admissions = n_distinct(APCS_Ident)) %>% 
  mutate(Admission_Date = as.Date(Admission_Date)) %>% 
  filter(Admission_Date > 2018-01-01) 
# Outpatient
b_hip_opa_daily <-
  a.opa_hip %>% 
  group_by(Appointment_Date) %>% 
  summarise(n_appointments = n_distinct(OPA_Ident)) %>% 
  mutate(Appointment_Date = as.Date(Appointment_Date)) %>% 
  filter(Appointment_Date > 2018-01-01) 
#ED
b_hip_ed_daily <-
  a.ed_hip %>% 
  group_by(Arrival_Date) %>% 
  summarise(n_attendances = n_distinct(EC_Ident)) %>% 
  mutate(Arrival_Date = as.Date(Arrival_Date)) %>% 
  filter(Arrival_Date > 2018-01-01)
#Merge
b_hip_all_daily <-
  b_hip_ip_daily %>%
  rename(Date = 1) %>% 
  left_join(b_hip_opa_daily %>% rename(Date = 1), by = ("Date")) %>% 
  left_join(b_hip_ed_daily %>% rename(Date = 1), by = ("Date")) %>% 
  pivot_longer(cols = -Date, names_to = "Activity", values_to = "Count") %>% 
  mutate(Activity = case_when(Activity == "n_admissions" ~ "Inpatient",
                              Activity == "n_appointments" ~ "Outpatient",
                              Activity == "n_attendances" ~ "Emergency Care",
                              TRUE ~ "NA")) %>% 
  mutate(Activity = factor(Activity, levels = c("Inpatient", "Outpatient", "Emergency Care")))


## B.1.2. Visualise Daily time series ####
# All hip activity  
b_hip_all_daily_vis <-
b_hip_all_daily %>% 
  filter(Date > "2018-01-01" & 
           Date < "2020-12-31") %>% 
  mutate(week_day = weekdays(Date)) %>% 
  mutate(weekend_flag = case_when(substr(week_day,1,1) == "S" ~ "Weekend",
                                  TRUE ~ "Week day")) %>% 
  ggplot(aes(x = Date, y = Count)) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = 0, ymax = 600, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = 0, ymax = 600, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  #geom_point(aes(colour = weekend_flag), size = 1, alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.2) +
  facet_wrap(~Activity) +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73")) +
  scale_color_SU(palette = "main") +
  labs(title = "Planned hip replacement activity was halted by the pandemic",
       subtitle = "Hip replacement procedures and related activity: National, planned and unplanned care",
       x = "Date",
       y = "Activity",
       colour = "")

## B.2. Interactive graphs #### 
# Plot multiple trends alongside each other
b_hip_all_month_wide <-
b_hip_all_month %>% 
  pivot_wider(id_cols = Month, names_from = Activity, values_from = Count) %>% clean_names()

ip <- xts(b_hip_all_month_wide$inpatient, order.by = b_hip_all_month_wide$month)
opa <- xts(b_hip_all_month_wide$outpatient, order.by = b_hip_all_month_wide$month)
ed <- xts(b_hip_all_month_wide$emergency_care, order.by = b_hip_all_month_wide$month)

hip_all <-
  cbind(ip, opa, ed)

interactive_dygraph_all_hip_activity <-
dygraph(hip_all, 
        main = "Hip replacement activity by care setting, England 2018-20", 
        xlab = "Date",
        ylab = "Monthly activity count") %>%
  dyOptions(drawGrid = FALSE, 
            colors = c("#f9bf07", "#5881c1", "#ec6555"),
            strokeWidth = 5) %>%
  dyCrosshair(direction = "vertical") %>%
  dyHighlight(highlightCircleSize = 5, 
              highlightSeriesBackgroundAlpha = 1, 
              hideOnMouseOut = FALSE) %>% 
  dyShading(from = "2020-03-23", to = "2020-07-04", color = "#d9d9d9") %>% 
  dyShading(from = "2020-10-1", to = "2021-01-04", color = "#d9d9d9") %>% 
  dyAnnotation("2020-05-13", text = "National Lockdown") %>% 
  dyAnnotation("2020-11-24", text = "3 Tier lockdown") 
## C. Cohort characteristics ####

## C.1. Age ####
c_ip_hip_age <-
a.ip_hip %>% 
  group_by(Admission_Date, Age_range) %>% 
  summarise(Admissions = n_distinct(APCS_Ident)) %>% 
  mutate(daily_sum = sum(Admissions)) %>% 
  mutate(prop = Admissions/daily_sum*100) %>% 
  mutate(prop = round(prop, 2)) %>% 
  ungroup()

c_ip_hip_age_vis <-
  c_ip_hip_age %>% 
  filter(Age_range %in% c("60-79", "80-99")) %>% 
  mutate(Admission_Date = as.Date(Admission_Date)) %>% 

  ggplot(aes(x = Admission_Date, y = prop, group = Age_range)) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = 0, ymax = 100, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = 0, ymax = 100, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  geom_point(aes(colour = Age_range), alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.2) +
  scale_color_SU(palette = "main") +
  labs(title = "The pandemic brought about a sharp increase in the rate of elderly hip procedures",
       subtitle = "Daily proportion of hip procedures performed by age range: National, elective inpatient episodes",
       x = "Date",
       y = "Proportion admissions",
       colour = "")

# Stacked bar chart 
# Stacked bar chart 
c_ip_hip_age_stacked_data <-
  c_ip_hip_age %>% 
  mutate(Admission_week = floor_date(as.Date(Admission_Date), unit = "week")) %>% 
  group_by(Admission_week, Age_range) %>% 
  summarise(sum = sum(Admissions)) %>% 
  mutate(prop = sum/sum(sum)*100) %>% 
  mutate(Proportion = round(prop, digits = 2)) %>% 
  drop_na(Proportion) 

c_ip_hip_age_stacked_vis <-
  c_ip_hip_age_stacked_data %>% 
  ggplot(aes(x = Admission_week, y = Proportion, fill = Age_range)) +
  geom_bar(position = "stack", stat = "identity", width = 7) +
  scale_fill_brewer(palette = "Spectral", limits=c("0-19", "20-39", "40-59", "60-79", "80-99", "100+")) +
  labs(title = "Proportion of elective hip procedures by age range",
       x = "Admission week",
       y = "Proportion",
       fill = "")


# Admissions in all ages as reference
c_ip_hip_admission_volume <-
a.ip_hip %>% 
  mutate(Admission_week = floor_date(as.Date(Admission_Date), unit = "week")) %>% 
  group_by(Admission_week) %>% 
  summarise(Admission_count = n()) %>% 
  mutate(ID = "Inpatient") 

c_ip_hip_admission_volume_vis <- 
  c_ip_hip_admission_volume %>%     
  ggplot(aes(x = Admission_week, y = Admission_count, colour = ID)) +
  geom_area(fill="#f9bf07", alpha=0.5) +
  geom_line(color="#f9bf07", size = 0.91) +
  geom_vline(xintercept = as.Date("2020-03-23"), linetype = "dashed") +
  geom_vline(xintercept = as.Date("2020-07-04"), linetype = "dashed") +
  annotate("text", x = as.Date("2020-05-13"), y = 2500, label = str_wrap("National lockdown", 10) ,size = 3) +
  geom_vline(xintercept = as.Date("2020-10-14"), linetype = "dashed") +
  geom_vline(xintercept = as.Date("2021-01-04"), linetype = "dashed") +
  annotate("text", x = as.Date("2020-11-24"), y = 2500, label = str_wrap("3 Tier lockdown", 10) ,size = 3) +
  annotate("text", x = as.Date("2020-08-24"), y = 2500, label = str_wrap("Recovery period", 10) ,size = 3) +
  labs(x = "Admission week", y = "Count", colour = "",
       title = "Count of weekly admissions for elective hip procedures")

# Bar and line chart - flat
(c_ip_hip_age_stacked_vis +
    geom_vline(xintercept = 18344, linetype = "dashed") +
    geom_vline(xintercept = 18447, linetype = "dashed") +
    geom_vline(xintercept = 18549, linetype = "dashed") +
    geom_vline(xintercept = 18631, linetype = "dashed")) /
  (c_ip_hip_admission_volume_vis 
   ) +
  plot_layout(heights = c(5,1)) 
  
# Combine interactive stacked bar and line graph 
c_ip_hip_stack_and_line <-
subplot(
  # Stacked bar 
ggplotly(c_ip_hip_age_stacked_vis +
           annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), y = 0, yend = 100, lty = 2) +
           annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), y = 0, yend = 100, lty = 2) +
           annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), y = 0, yend = 100, lty = 2) +
           annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), y = 0, yend = 100, lty = 2),
         tooltip = c("Age_range", "Proportion")
         ),

  # Line chart
style(
  ggplotly(c_ip_hip_admission_volume_vis +
             annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), y = 0, yend = 2800, lty = 2) +
             annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), y = 0, yend = 2800, lty = 2) +
             annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), y = 0, yend = 2800, lty = 2) +
             annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), y = 0, yend = 2800, lty = 2),
           tooltip = c("Admission_week", "Admission_count")),
  showlegend = FALSE
  ),
  # Organise plots
nrows = 2, heights = c(0.8,0.2), shareY = TRUE, shareX = TRUE 
      ) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = paste0('<b> At the height of disruption, care was limited to the older patients despited increased covid risk <b>',
                                    '<br>',
                                    '<sup>',
                                    '<i>Proportion and count of elective hip procedures by age range: National, weekly average<i>',
                                    '</sup>')))   
## C.2. Deprivation #### 
con_ref <- dbConnect(odbc(), 
                 Driver = "SQL Server", 
                 server = "PRODNHSESQL101", 
                 Database = "NHSE_Reference")

c_lsoa_imd <- tbl(con_ref, "tbl_Ref_Other_Deprivation_By_LSOA") %>% 
  collect()

# Disconnect! # 
dbDisconnect(con_ref) 

c_ip_hip_imd <-
a.ip_hip  %>% 
  select(1:9, procedure_type, Age_range) %>% 
  left_join(select(c_lsoa_imd, LSOA_Code, IMD_Score, IMD_Rank, IMD_Decile), by = c("Der_Postcode_LSOA_2011_Code" = "LSOA_Code")) %>% 
  mutate(IMD_Quintile = case_when(IMD_Decile %in% c(1,2) ~ 1,
                                  IMD_Decile %in% c(3,4) ~ 2,
                                  IMD_Decile %in% c(5,6) ~ 3,
                                  IMD_Decile %in% c(7,8) ~ 4,
                                  IMD_Decile %in% c(9,10) ~ 5,
                                  TRUE ~ 99)) %>% 
  group_by(Admission_Date, IMD_Quintile) %>% 
  summarise(Admissions = n_distinct(APCS_Ident)) %>% 
  mutate(daily_sum = sum(Admissions)) %>% 
  mutate(prop = Admissions/daily_sum*100) %>% 
  mutate(prop = round(prop, 2)) 

c_ip_hip_imd_wk <-
c_ip_hip_imd %>% 
  ungroup() %>% 
  mutate(Admission_week = floor_date(as.Date(Admission_Date), unit = "week")) %>% 
  group_by(Admission_week, IMD_Quintile) %>% 
  summarise(sum = sum(Admissions)) %>% 
  mutate(prop = sum/sum(sum)*100) %>% 
  mutate(IMD_Quintile = as.character(IMD_Quintile))

c.ip_hip_imd_vis <-
  c_ip_hip_imd_wk %>% 
  ggplot(aes(x = Admission_week, y = prop, group = IMD_Quintile, colour = IMD_Quintile  )) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = 0, ymax = 40, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = 0, ymax = 40, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  annotate("text", x = as.Date("2020-05-13"), y = 37, label = "National lockdown",size = 3, angle = 90) +
  annotate("text", x = as.Date("2020-11-24"), y = 34, label = "3 Tier lockdown",size = 3, angle = 90) +
  geom_point( alpha = 0.75) +
  geom_smooth(method = "loess", span = 0.2) +
  scale_colour_brewer(palette = "Spectral") +
  ylim(0,40) +
  labs(title = "During the initial lockown, hip care was focused on the most and least deprived",
       subtitle = "Weekly proportion of hip procedures performed by IMD deprivation quintile: National, elective inpatient episodes",
       x = "Admission week",
       y = "Proportion admissions",
       caption = "Note: IMD quintile 1 represents the most deprived 20% of the population",
       colour = "")

# Total stacked bar 
c_ip_hip_imd_stack_vis <-
c_ip_hip_imd %>%  
  mutate(Admission_week = floor_date(as.Date(Admission_Date), unit = "week")) %>% 
  group_by(Admission_week, IMD_Quintile) %>% 
  summarise(sum = sum(Admissions)) %>% 
  mutate(prop = sum/sum(sum)*100) %>% 
  mutate(Proportion = round(prop, digits = 2)) %>% 
  drop_na(Proportion) %>% 
  
  ggplot(aes(x = Admission_week, y = Proportion, fill = as.character(IMD_Quintile))) +
  geom_bar(position = "stack", stat = "identity", width = 7) +
  scale_fill_brewer(palette = "Spectral") +
  labs(title = "Proportion of elective hip procedures by IMD Quintile",
       subtitle = "National, weekly average", 
       x = "Admission week",
       y = "Proportion",
       fill = "")

# Stacked bar and line graph
c_hip_ip_stacked_bar_and_line <-
subplot(
  # Stacked bar 
  ggplotly(c_ip_hip_imd_stack_vis +
             annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), y = 0, yend = 100, lty = 2) +
             annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), y = 0, yend = 100, lty = 2) +
             annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), y = 0, yend = 100, lty = 2) +
             annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), y = 0, yend = 100, lty = 2),
           tooltip = c("Admission_week","Proportion")
  ),
  
  # Line chart
  style(
    ggplotly(c_ip_hip_admission_volume_vis +
               annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), y = 0, yend = 2800, lty = 2) +
               annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), y = 0, yend = 2800, lty = 2) +
               annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), y = 0, yend = 2800, lty = 2) +
               annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), y = 0, yend = 2800, lty = 2),
             tooltip = c("Admission_week", "Admission_count")),
    showlegend = FALSE
  ),
  # Organise plots
  nrows = 2, heights = c(0.8,0.2), shareY = TRUE, shareX = TRUE 
) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = paste0('<b>More affluent people received a greater than average proportion of hip care during lockdown<b>',
                                    '<br>',
                                    '<sup>',
                                    '<i>Proportion and count of elective hip procedures by deprivation quintile: National, weekly average<i>',
                                    '</sup>'))) 


## C.2.2. Deprivation x Age ####
c_ip_hip_imd_age_wk <-
  a.ip_hip  %>% 
  select(1:9, procedure_type, Age_range) %>% 
  left_join(select(c_lsoa_imd, LSOA_Code, IMD_Score, IMD_Rank, IMD_Decile), by = c("Der_Postcode_LSOA_2011_Code" = "LSOA_Code")) %>% 
  mutate(IMD_Quintile = case_when(IMD_Decile %in% c(1,2) ~ 1,
                                  IMD_Decile %in% c(3,4) ~ 2,
                                  IMD_Decile %in% c(5,6) ~ 3,
                                  IMD_Decile %in% c(7,8) ~ 4,
                                  IMD_Decile %in% c(9,10) ~ 5,
                                  TRUE ~ 99)) %>% 
  mutate(Admission_week = floor_date(as.Date(Admission_Date), unit = "week")) %>% 
  group_by(Admission_week, Age_range, IMD_Quintile) %>% 
  summarise(Admissions = n_distinct(APCS_Ident)) %>%
  ungroup() %>% group_by(Admission_week) %>% 
  mutate(weekly_sum = sum(Admissions)) %>% 
  mutate(Proportion = Admissions/weekly_sum*100) %>% 
  mutate(Proportion = round(Proportion, 2)) 

c_ip_hip_imd_age_wk_vis <-
c_ip_hip_imd_age_wk %>% 
  filter(Age_range %in% c("40-59", "60-79", "80-99")) %>% 
  drop_na(Age_range) %>% 
  filter(substr(Admission_week,1,4) > 2018) %>% 
  ggplot(aes(x = Admission_week, y = Proportion, 
             colour = as.character(IMD_Quintile), group = as.character(IMD_Quintile))) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = 0, ymax = 20, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = 0, ymax = 20, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  geom_point( alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.2) +
  scale_color_SU(palette = "main") +
  facet_grid(IMD_Quintile ~ Age_range, scale = "free_y") +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73"),
        legend.position = "none",
        axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) +
  labs(title = "Deprivation has distinct influences on care access, independent of age range",
       subtitle = "Weekly proportion of hip procedures performed by age range and deprivation: National, elective inpatient episodes",
       x = "Admission week",
       y = "Proportion admissions",
       colour = "")

## C.3. Ethnicity ####
c_ethnicity_lookup <-
tribble(
  ~Ethnic_Category, ~Ethnic_Category_Desc, ~Ethnicity_broad,
 
  "99",	"Not known", "Not known",	
  "A",	"British", "White",
  "B",	"Irish"	, "White",
  "C",	"Any other white background", "White",
  "D",	"White and black caribbean",	"Mixed",
  "E",	"White and black african",	"Mixed",
  "F",	"White and asian",	"Mixed",
  "G",	"Any other mixed background", "Mixed",
  "H",	"Indian", "Asian",
  "J",	"Pakistani",	"Asian",
  "K",	"Bangladeshi",	"Asian",
  "L",	"Any other asian background", "Asian",
  "M",	"Caribbean",	"Black",
  "N",	"African", "Black",
  "P",	"Any other black background", "Black",
  "R",	"Chinese", "Asian",
  "S",	"Any other ethnic group", "Other",
  "Z",	"Not stated", "Not known",
  "NA", "Not known", "Not known"
  )

## C.3.1. Full ethnicity ####
c_ip_hip_ethnicity <-
  a.ip_hip %>% 
  mutate(Ethnic_Group = case_when(Ethnic_Group %in% c("99", "NA") ~ Ethnic_Group,
                                  TRUE ~ substr(Ethnic_Group,1,1)
                                  )) %>% # Shorten Ethnic group to first character to clean data (except when coded as 99 or NA)
  mutate(Ethnic_Group = case_when(is.na(Ethnic_Group) ~ "99",
                                  TRUE ~ Ethnic_Group)) %>% # Replace NA's with "NA" to left_join to "Not Known"
  left_join(c_ethnicity_lookup, by = c("Ethnic_Group" = "Ethnic_Category")) %>% 
  group_by(Admission_Date, Ethnic_Group, Ethnic_Category_Desc, Ethnicity_broad) %>% 
  summarise(Admissions = n_distinct(APCS_Ident)) %>% 
  ungroup() %>%  group_by(Admission_Date) %>% 
  mutate(daily_sum = sum(Admissions)) %>% 
  mutate(prop = Admissions/daily_sum*100)

# Weekly ethnicity proportions 
c_ip_hip_ethnicity_wk <-
  c_ip_hip_ethnicity %>% 
  ungroup() %>% 
  mutate(Admission_week = floor_date(as.Date(Admission_Date), unit = "week")) %>% 
  group_by(Admission_week, Ethnic_Category_Desc) %>% 
  summarise(sum = sum(Admissions)) %>% 
  mutate(prop = sum/sum(sum)*100) %>% 
  mutate(Proportion = round(prop, digits = 2)) %>% 
  drop_na(Proportion) %>% 
  mutate(Ethnicity = Ethnic_Category_Desc)

c_ip_hip_ethnicity_wk_vis <-
c_ip_hip_ethnicity_wk %>% 
  ggplot(aes(x = Admission_week, y = Proportion, group = Ethnic_Category_Desc, colour = Ethnicity )) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = 0, ymax = 100, fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = 0, ymax = 100, fill= "#fed976", alpha = 0.8) + # 3 tier system
  annotate("text", x = as.Date("2020-05-13"), y = 45, label = "National lockdown",size = 3, angle = 90) +
  annotate("text", x = as.Date("2020-11-24"), y = 40, label = "3 Tier lockdown",size = 3, angle = 90) +
  geom_point( alpha = 0.75) +
  geom_smooth(method = "loess", span = 0.2) +
  scale_color_SU(palette = "main") +
  labs(title = "Weekly proportion of hip procedures performed by ethnicity",
       subtitle = "National, elective inpatient episodes",
       x = "Admission week",
       y = "Proportion admissions",
       colour = "Ethnicity")

# Facet wrap
c_ip_hip_ethnicity_wk_vis_facet <-
  c_ip_hip_ethnicity_wk %>% 
  left_join(c_ethnicity_lookup, by = c("Ethnic_Category_Desc")) %>% 
  mutate(
    Ethnicity = factor(
      Ethnicity,
        levels = c(
          # Asian
          "Bangladeshi", "Chinese", "Indian", "Pakistani", "Any other asian background",
          # Black
          "African", "Caribbean", "Any other black background",
          # Mixed
          "White and asian", "White and black african", "White and black caribbean", "Any other mixed background",
          # Not known
          "Not known", "Not stated", "Any other ethnic group",
          # White
          "British", "Irish", "Any other white background"
          ))) %>% 
  ggplot(aes(x = Admission_week, y = Proportion,  colour = Ethnicity_broad )) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = -Inf, ymax = Inf, fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = -Inf, ymax = Inf, fill= "#fed976", alpha = 0.8) + # 3 tier system
  geom_point( alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.2) +
  facet_wrap(~Ethnicity, scales = "free") +
    theme(strip.text = element_text(size = 6.5, face = "bold"),
          #strip.background = element_rect(colour = "#686f73"),
          legend.position = c(0.8, 0.09)
          ) +
  guides(col = guide_legend(ncol = 2)) +
  scale_color_SU(palette = "main") +
  labs(title = "Weekly proportion of hip procedures performed by ethnicity",
       subtitle = "National, elective inpatient episodes",
       x = "Admission week",
       y = "Proportion admissions",
       colour = "Broad ethnicity")


# Stacked bar chart (2018 & 2019 annual averages vs 2020 weekly averages) 
# Full ethnicity descriptions
c_ip_hip_ethnicity_stack_vis <-
c_ip_hip_ethnicity_wk %>% 
  
  ggplot(aes(x = Admission_week, y = Proportion, fill = Ethnicity)) +
  geom_bar(position = "stack", stat = "identity", width = 7) +
  scale_fill_SU(palette = "mixed") +
  theme(axis.text.y = element_blank()) +
  labs(title = "Proportion of elective hip procedures by ethnicity",
       subtitle = "National, weekly average", 
       x = "Admission week",
       y = "Proportion",
       fill = "")


## C.3.2. Broad ethnicity ####
c_ip_hip_ethnicity_broad <-
c_ip_hip_ethnicity %>% 
  ungroup() %>% 
  mutate(Admission_week = floor_date(as.Date(Admission_Date), unit = "week")) %>% 
  group_by(Admission_week, Ethnicity_broad) %>% 
  summarise(sum = sum(Admissions)) %>% 
  mutate(prop = sum/sum(sum)*100) %>% 
  ungroup() %>% 
  mutate(Proportion = round(prop, digits = 2)) %>% 
  drop_na(Proportion)

c_ip_hip_ethnicity_broad_vis <-
c_ip_hip_ethnicity_broad %>% 
  filter(Admission_week < "2021-01-03") %>% 
  ggplot(aes(x = Admission_week, y = Proportion, group = Ethnicity_broad, colour = Ethnicity_broad  )) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = 0, ymax = 100, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = 0, ymax = 100, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  annotate("text", x = as.Date("2020-05-13"), y = 50, label = "National lockdown",size = 3, angle = 90) +
  annotate("text", x = as.Date("2020-11-24"), y = 50, label = "3 Tier lockdown",size = 3, angle = 90) +
  geom_point( alpha = 0.75) +
  geom_smooth(method = "loess", span = 0.2) +
  scale_color_SU(palette = "main") +
  labs(title = "Weekly proportion of hip procedures performed by ethnicity",
       subtitle = "National, elective inpatient episodes",
       x = "Admission week",
       y = "Proportion admissions",
       colour = "")

# Stacked bar chart 
c_ip_hip_ethnicity_broad_stack_vis <-
  c_ip_hip_ethnicity_broad %>%  
  ggplot(aes(x = Admission_week, y = Proportion, fill = Ethnicity_broad)) +
  geom_bar(position = "stack", stat = "identity", width = 7) +
  scale_fill_SU(palette = "main") +
  theme(axis.text.y = element_blank()) +
  labs(title = "Proportion of elective hip procedures by broad ethnicity",
       subtitle = "National, weekly average", 
       x = "Admission week",
       y = "Proportion",
       fill = "")

# Stacked bar and line graph
c_hip_ip_stacked_bar_and_line_ethnicity <-
  subplot(
    # Stacked bar 
    ggplotly(c_ip_hip_ethnicity_broad_stack_vis +
               annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), 
                        y = 0, yend = 100, lty = 2) +
               annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), 
                        y = 0, yend = 100, lty = 2) +
               annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), 
                        y = 0, yend = 100, lty = 2) +
               annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), 
                        y = 0, yend = 100, lty = 2),
             tooltip = c("Admission_week","Proportion")
    ),
    
    # Line chart
    style(
      ggplotly(c_ip_hip_admission_volume_vis +
                 annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), 
                          y = 0, yend = 2800, lty = 2) +
                 annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), 
                          y = 0, yend = 2800, lty = 2) +
                 annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), 
                          y = 0, yend = 2800, lty = 2) +
                 annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), 
                          y = 0, yend = 2800, lty = 2),
               tooltip = c("Admission_week", "Admission_count")),
      showlegend = FALSE
    ),
    # Organise plots
    nrows = 2, heights = c(0.8,0.2), shareY = TRUE, shareX = TRUE 
  ) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = paste0('<b>Proportion and count of elective hip procedures by broad ethnicity<b>',
                                    '<br>',
                                    '<sup>',
                                    '<i>National, weekly average<i>',
                                    '</sup>'))) 

#Facet wrap broad ethnicities 
c_ip_hip_ethnicity_broad_facet <-
c_ip_hip_ethnicity_broad %>%
  drop_na(Proportion) %>% 
  filter(Proportion != 0) %>% 
  filter(Admission_week < "2021-01-03") %>% 
  ggplot(aes(x = Admission_week, y = Proportion, group = Ethnicity_broad, colour = Ethnicity_broad  )) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = -Inf, ymax = Inf, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = -Inf, ymax = Inf, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  geom_point( alpha = 0.75) +
  geom_smooth(method = "loess", span = 0.2) +
  scale_color_SU(palette = "main") +
  facet_wrap(~Ethnicity_broad, scales = "free") +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73"),
        legend.position = "none") +
  labs(title = "Weekly proportion of hip procedures performed by ethnicity",
       subtitle = "National, elective inpatient episodes",
       x = "Admission week",
       y = "Proportion admissions",
       colour = "")

## E.3. Admission deficit #### 
e_hip_ip_week_number <-
  b_hip_ip_daily %>% 
  mutate(year = year(Admission_Date),
         month = months(Admission_Date), 
         week_number = isoweek(Admission_Date)) %>% 
  group_by(year, week_number) %>% 
  summarise(weekly_admissions = sum(n_admissions)) %>% 
  ungroup()

e_hip_ip_week_number_wide <-
e_hip_ip_week_number %>% 
  filter(year == 2020) %>% 
  rename(n_2020 = weekly_admissions) %>% 
  select(-year) %>% 
  left_join(e_hip_ip_week_number %>% 
              filter(year == 2018) %>% 
              select(-year) %>% 
              rename(n_2018 = weekly_admissions), by = c("week_number"), keep = FALSE) %>% 
  left_join(e_hip_ip_week_number %>% 
              filter(year == 2019) %>% 
              select(-year) %>% 
              rename(n_2019 = weekly_admissions), by = c("week_number"), keep = FALSE) %>% 
  group_by(week_number) %>% 
  mutate(Avg1819 =  round(mean(c(n_2018, n_2019), na.rm=T))) %>% 
  ungroup() %>% 
  mutate(def = Avg1819 - n_2020) %>% 
  mutate(PC_def = case_when(
    week_number < 13 ~ 0,
    week_number >= 13 ~ Avg1819 - n_2020)) # Additional variable to start deficit from lockdown start

e_hip_ip_annual_trends <-
e_hip_ip_week_number_wide %>%   
  ggplot(aes(x = week_number, y = n_2020)) +
  geom_smooth(method = "loess", span = 0.2, colour = "#f9bf07") +
  geom_smooth(aes(y = Avg1819), method = "loess", span = 0.2, colour = "#5881c1") +
  # Covid timeline markers
  annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
  annotate("text", x = 20, y = 3250, label = str_wrap("National lockdown", 10) ,size = 3.5) +
  annotate("text", x = 34, y = 3250, label = str_wrap("Recovery period", 10) ,size = 3.5) +
  annotate("text", x = 48, y = 3250, label = str_wrap("3 Tier lockdown", 10) ,size = 3.5) +
  # Labels and arrows 
  annotate(geom = "curve", x = 23, y = 1725, xend = 23, yend = 2250, 
           curvature = 0.3, arrow = arrow(length = unit(2, "mm"))) +
  annotate(geom = "curve", x = 35, y = 500, xend = 33, yend = 850, 
          curvature = -0.3, arrow = arrow(length = unit(2, "mm"))) +
  annotate("text", x = 18, y = 1725, label = str_wrap("2018-19 average",10), size = 3) +
  annotate("text", x = 38, y = 500, label = "2020", size = 3) +
  labs(x = "Week number", y = "Admissions")

# Visualise admission deficit 
e_hip_ip_admission_def_vis <-
e_hip_ip_week_number_wide %>% 
  ggplot(aes(x = week_number, y = PC_def)) +
  geom_smooth(method = "loess", span = 0.2, colour = "#f9bf07") +
  # Covid timeline markers
  annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
  annotate("text", x = 20, y = 3250, label = str_wrap("National lockdown", 10) ,size = 3.5) +
  annotate("text", x = 34, y = 3250, label = str_wrap("Recovery period", 10) ,size = 3.5) +
  annotate("text", x = 48, y = 3250, label = str_wrap("3 Tier lockdown", 10) ,size = 3.5) +
  # Labels and arrows 
  annotate(geom = "curve", x = 33, y = 500, xend = 34, yend = 850, 
           curvature = 0.3, arrow = arrow(length = unit(2, "mm"))) +
  annotate("text", x = 31, y = 500, label = "2020", size = 3) +
  labs(x = "Week number", y = "Admission deficit")

# Patch
e_hip_ip_annual_trends + e_hip_ip_admission_def_vis + 
  plot_annotation(title = "Yearly patterns in hip replacement admissions",
                        subtitle = "National, Weekly elective inpatient procedures, 2018-20"
                        )

# Total admission deficit number to quote
e_hip_ip_total_admission_def <-
e_hip_ip_week_number_wide %>% 
  select(PC_def) %>% 
  sum(na.rm = TRUE)

## E.3.2. Demographic index: Comparing 2020 to 18/19 mean by subgroups ####
# Admission count by demographic subgroups  
e_hip_ip_demographic_index <-
a.ip_hip %>% 
  #Ethnicity
  mutate(
    Ethnic_Group = case_when(Ethnic_Group %in% c("99", "NA") ~ Ethnic_Group,
                             TRUE ~ substr(Ethnic_Group,1,1)
                             )) %>% # Shorten Ethnic group to first character to clean data (except when coded as 99 or NA)
  mutate(
    Ethnic_Group = case_when(is.na(Ethnic_Group) ~ "99",
                             TRUE ~ Ethnic_Group)) %>% # Replace NA's with "NA" to left_join to "Not Known"
  left_join(c_ethnicity_lookup, by = c("Ethnic_Group" = "Ethnic_Category")) %>% 
  # Deprivation 
  left_join(select(c_lsoa_imd, LSOA_Code, IMD_Score, IMD_Rank, IMD_Decile), by = c("Der_Postcode_LSOA_2011_Code" = "LSOA_Code")) %>% 
  mutate(IMD_Quintile = case_when(IMD_Decile %in% c(1,2) ~ 1,
                                  IMD_Decile %in% c(3,4) ~ 2,
                                  IMD_Decile %in% c(5,6) ~ 3,
                                  IMD_Decile %in% c(7,8) ~ 4,
                                  IMD_Decile %in% c(9,10) ~ 5,
                                  TRUE ~ 99)) %>% 
  # Activity count by demographic and region
  mutate(Admission_Date = as.Date(Admission_Date)) %>% 
  group_by(Admission_Date, nhser20nm, Age_range, Sex, Ethnic_Category_Desc, Ethnicity_broad, IMD_Quintile) %>% 
  summarise(Admission_count = n_distinct(APCS_Ident)) %>% 
  ungroup()


# Function: get admission count and deficit by demographic 
e_get_admission_deficit <- function(demographic_var) {
  
  e_hip_ip_demographic_index <- 
    e_hip_ip_demographic_index %>%
    mutate(demographic_var = {{demographic_var}})
  
  sub_data <-
  e_hip_ip_demographic_index %>% 
    mutate(year = year(Admission_Date),
           month = months(Admission_Date), 
           week_number = isoweek(Admission_Date)) %>% 
    group_by(year, week_number, demographic_var) %>% 
    summarise(weekly_admissions = sum(Admission_count)) %>% 
    ungroup()
  
  sub_data_wide <-
    sub_data %>% 
    filter(year == 2020) %>% 
    rename(n_2020 = weekly_admissions) %>% 
    select(-year) %>% 
    left_join(sub_data %>% 
                filter(year == 2018) %>% 
                select(-year) %>% 
                rename(n_2018 = weekly_admissions), by = c("week_number", 'demographic_var'), keep = FALSE) %>% 
    left_join(sub_data %>% 
                filter(year == 2019) %>% 
                select(-year) %>% 
                rename(n_2019 = weekly_admissions), by = c("week_number", 'demographic_var'), keep = FALSE) %>% 
    group_by(week_number, demographic_var) %>% 
    mutate(Avg1819 =  round(mean(c(n_2018, n_2019), na.rm=T))) %>% 
    mutate(def = Avg1819 - n_2020) %>% 
    mutate(PC_def = case_when(
      week_number < 13 ~ 0,
      week_number >= 13 ~ Avg1819 - n_2020)) %>%  # Additional variable to start deficit from lockdown start
    ungroup() %>% 
    filter(week_number != 53)
  
  sub_data_wide
}

# Function: Plot 2020 vs 2018/19 average, facet by demographic
facet_plot_function <- function(data) {
  
  data %>% 
    pivot_longer(cols = c(-week_number, - demographic_var),
                 names_to = "trend", 
                 values_to = "Admission_count") %>% 
    filter(trend %in% c("n_2020", "Avg1819")) %>% 
    mutate(trend = case_when(trend == "n_2020" ~ "2020",
                             trend == "Avg1819" ~ "2018-19"))  %>% 
    
    ggplot(aes(x = week_number, y = Admission_count, colour = trend, group = trend)) +
    geom_smooth(method = "loess", span = 0.2) +
    scale_colour_manual(values = c("#f9bf07", "#5881c1" )) +
    facet_wrap(~ demographic_var) +
    # Covid timeline markers
    annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
    annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
    annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
    annotate("text", x = 20, y = -150, label = str_wrap("Lockdown", 10) ,size = 2.5) +
    annotate("text", x = 34, y = -150, label = str_wrap("Recovery", 10) ,size = 2.5) +
    annotate("text", x = 48, y = -150, label = str_wrap("Lockdown", 10) ,size = 2.5) +
    theme(strip.text = element_text(size = 10, face = "bold"),
          legend.text = element_text(size = 12)) 
  }

# Total Admission deficit by region - Table 
e_hip_ip_total_deficit_region <-
  e_get_admission_deficit(nhser20nm) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var) %>% 
  mutate(LCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100,2))

# Cumulative admission deficit (from lockdown 1) 
e_weekly_cum_def_by_region <-
e_get_admission_deficit(nhser20nm) %>% 
  group_by(demographic_var) %>% 
  mutate(cum_def = cumsum(PC_def)) %>% 
  pivot_longer(cols = c(-week_number, -demographic_var),
               names_to = "trend", 
               values_to = "count") %>% 
  filter(trend == "cum_def") 

# Deprivation
e_hip_ip_def_deprivation <-
e_get_admission_deficit(IMD_Quintile) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var) %>% 
  mutate(LCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100,2))  

e_hip_ip_def_deprivation_change <-
e_get_admission_deficit(IMD_Quintile) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_2020 = sum(n_2020),
            sum_1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(prop_2020 = round(sum_2020/sum(sum_2020)*100,2),
         prop_1819 = round(sum_1819/sum(sum_1819)*100,2),
         prop_change = prop_2020 - prop_1819) 

# Broad ethnicity
e_hip_ip_def_ethnicity_broad <-
e_get_admission_deficit(Ethnicity_broad) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = sum_def/sum_Avg1819*100) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var) %>% 
  mutate(LCI = add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,
         UCI = add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100) 

# Ethnicity
e_hip_ip_def_ethnicity <-
e_get_admission_deficit(Ethnic_Category_Desc) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = sum_def/sum_Avg1819*100) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var) %>% 
  filter(def_prop > 0) %>% 
  mutate(LCI = add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,
         UCI = add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100) 

e_hip_ip_def_ethnicity_prop_change <-
e_get_admission_deficit(Ethnic_Category_Desc) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_2020 = sum(n_2020),
            sum_1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(prop_2020 = sum_2020/sum(sum_2020)*100,
         prop_1819 = sum_1819/sum(sum_1819)*100,
         prop_change = prop_2020 - prop_1819) %>%
  mutate(demographic_var = fct_reorder(demographic_var, prop_change)) 

# Age range
e_hip_ip_def_age_range <-
e_get_admission_deficit(Age_range) %>% 
  drop_na(demographic_var) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = sum_def/sum_Avg1819*100) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var) %>% 
  filter(def_prop > 0) %>% 
  mutate(LCI = add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,
         UCI = add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100) 

## E 3.3. Comparing admissions by demographic and regions ####
# Function: get admission count and deficit by demographic and region
e_get_admission_deficit_mult <- function(demographic_var1, demographic_var2) {
  
  e_hip_ip_demographic_index <- 
    e_hip_ip_demographic_index %>%
    mutate(demographic_var1 = {{demographic_var1}},
           demographic_var2 = {{demographic_var2}})
  
  sub_data <-
    e_hip_ip_demographic_index %>% 
    mutate(year = year(Admission_Date),
           month = months(Admission_Date), 
           week_number = isoweek(Admission_Date)) %>% 
    group_by(year, week_number, demographic_var1, demographic_var2) %>% 
    summarise(weekly_admissions = sum(Admission_count)) %>% 
    ungroup()
  
  sub_data_wide <-
    sub_data %>% 
    filter(year == 2020) %>% 
    rename(n_2020 = weekly_admissions) %>% 
    select(-year) %>% 
    left_join(sub_data %>% 
                filter(year == 2018) %>% 
                select(-year) %>% 
                rename(n_2018 = weekly_admissions), by = c("week_number", "demographic_var1", "demographic_var2"), keep = FALSE) %>% 
    left_join(sub_data %>% 
                filter(year == 2019) %>% 
                select(-year) %>% 
                rename(n_2019 = weekly_admissions), by = c("week_number", "demographic_var1", "demographic_var2"), keep = FALSE) %>% 
    group_by(week_number, demographic_var1, demographic_var2) %>% 
    mutate(Avg1819 =  round(mean(c(n_2018, n_2019), na.rm=T))) %>% 
    mutate(def = Avg1819 - n_2020) %>% 
    mutate(PC_def = case_when(
      week_number < 13 ~ 0,
      week_number >= 13 ~ Avg1819 - n_2020)) %>%  # Additional variable to start deficit from lockdown start
    
    ungroup() 
  
  sub_data_wide 
}

# NHS region x ethincity - change in prop
e_deficit_nhsr_ethnicity_prop_change <-
e_get_admission_deficit_mult(nhser20nm, Ethnicity_broad) %>% 
  drop_na(demographic_var1, demographic_var2) %>% 
  group_by(demographic_var1, demographic_var2) %>% 
  summarise(sum_n_2020 = sum(n_2020, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(prop_2020 = sum_n_2020/sum(sum_n_2020)*100,
         prop_1819 = sum_Avg1819/sum(sum_Avg1819)*100,
         prop_change = prop_2020 - prop_1819)  # 2020-Avg201819


# NHSE region x Deprivation
e_hip_ip_nhserXdeprivation <-
e_get_admission_deficit_mult(nhser20nm, IMD_Quintile) %>% 
  drop_na(demographic_var1, demographic_var2) %>% 
  group_by(demographic_var1, demographic_var2) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
          sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var1, demographic_var2) %>% 
  filter(def_prop > 0) %>% 
  mutate(LCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100,2)) %>%
  arrange(demographic_var1, demographic_var2)


# Deprivation x Broad ethnicity
e_hip_ip_def_deprivationXEthnicity_broad <-
e_get_admission_deficit_mult(IMD_Quintile, Ethnicity_broad) %>% 
  drop_na(demographic_var1, demographic_var2) %>% 
  group_by(demographic_var1, demographic_var2) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>% 
  mutate(sum_def_adj = case_when(sum_def >= 0 ~ sum_def,
                                 sum_def < 0 ~ -sum_def )) %>% # Switch out the negative def to positive to calcuate CI
  group_by(demographic_var1, demographic_var2) %>% 
  #filter(def_prop > 0) %>% 
  mutate(LCI = round(add4ci(x = sum_def_adj, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def_adj, n = sum_Avg1819,0.95)$conf.int[2]*100,2)) %>%
  mutate(LCI = case_when(sum_def >= 0 ~ LCI, sum_def < 0 ~ -LCI),
         UCI = case_when(sum_def >= 0 ~ UCI, sum_def < 0 ~ -UCI)) %>%  # Return CI's to negative where the def was negative
  arrange(demographic_var2, demographic_var1)

# Deprivation x Ethnicity - Asian specific 
e_hip_ip_def_deprivationXEthnicity_broad_Asain <-
e_get_admission_deficit_mult(IMD_Quintile, Ethnic_Category_Desc ) %>% 
  left_join(c_ethnicity_lookup %>% 
              select(-Ethnic_Category), by = c("demographic_var2" = "Ethnic_Category_Desc")) %>% 
  filter(Ethnicity_broad == "Asian") %>% 
  drop_na(demographic_var1, demographic_var2) %>% 
  group_by(demographic_var1, demographic_var2) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>% 
  mutate(sum_def_adj = case_when(sum_def >= 0 ~ sum_def,
                                 sum_def < 0 ~ -sum_def )) %>% # Switch out the negative def to positive to calcuate CI
  group_by(demographic_var1, demographic_var2) %>% 
  #filter(def_prop > 0) %>% 
  mutate(LCI = round(add4ci(x = sum_def_adj, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def_adj, n = sum_Avg1819,0.95)$conf.int[2]*100,2)) %>%
  mutate(LCI = case_when(sum_def >= 0 ~ LCI, sum_def < 0 ~ -LCI),
         UCI = case_when(sum_def >= 0 ~ UCI, sum_def < 0 ~ -UCI))

e_hip_ip_def_deprivationXEthnicity_broad_change_prop <-
e_get_admission_deficit_mult(IMD_Quintile, Ethnicity_broad) %>% 
  drop_na(demographic_var1, demographic_var2) %>% 
  group_by(demographic_var1, demographic_var2) %>% 
  summarise(sum_n_2020 = sum(n_2020, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(prop_2020 = sum_n_2020/sum(sum_n_2020)*100,
         prop_1819 = sum_Avg1819/sum(sum_Avg1819)*100,
         prop_change = prop_2020 - prop_1819) 

# Deprivation x Age
e_hip_ip_def_deprivation_age <-
e_get_admission_deficit_mult(IMD_Quintile, Age_range) %>% 
  filter(demographic_var2 != "100+") %>% 
  drop_na(demographic_var1, demographic_var2) %>% 
  group_by(demographic_var1, demographic_var2) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE)) %>% 
  left_join(
    e_get_admission_deficit_mult(IMD_Quintile, Age_range) %>% 
      group_by(demographic_var1, demographic_var2) %>% 
      summarise(sum_n = sum(Avg1819, na.rm = TRUE)), 
    by = c("demographic_var1", "demographic_var2")) %>% 
  mutate(proportion = sum_def/sum_n*100) %>% 
  mutate(sum_def_adj = case_when(sum_def >= 0 ~ sum_def,
                                 sum_def < 0 ~ -sum_def )) %>% # Switch out the negative def to positive to calcuate CI
  group_by(demographic_var1, demographic_var2) %>% 
  mutate(LCI = add4ci(x = sum_def_adj, n = sum_n,0.95)$conf.int[1]*100,
         UCI = add4ci(x = sum_def_adj, n = sum_n,0.95)$conf.int[2]*100) %>%
  mutate(LCI = case_when(sum_def >= 0 ~ LCI, sum_def < 0 ~ -LCI),
         UCI = case_when(sum_def >= 0 ~ UCI, sum_def < 0 ~ -UCI)) %>%  # Return CI's to negative where the def was negative
  arrange(demographic_var2, demographic_var1) 


## E.4. Emergency admissions for hip replacement procedures ####
a.ip_hip_unplanned <-
  a.ip_hip_unplanned %>% 
  mutate(Age_range = case_when(Age_at_Start_of_Episode_SUS >= 0 & Age_at_Start_of_Episode_SUS < 20 ~ "0-19",
                               (Age_at_Start_of_Episode_SUS >= 20 & Age_at_Start_of_Episode_SUS <40) ~  "20-39",
                               (Age_at_Start_of_Episode_SUS >= 40 & Age_at_Start_of_Episode_SUS <60) ~  "40-59",
                               (Age_at_Start_of_Episode_SUS >= 60 & Age_at_Start_of_Episode_SUS <80) ~  "60-79",
                               (Age_at_Start_of_Episode_SUS >= 80 & Age_at_Start_of_Episode_SUS <100) ~ "80-99",
                               Age_at_Start_of_Episode_SUS >= 100 ~ "100+")) %>% 
  mutate(Age_range = factor(Age_range, levels = c("0-19", 
                                                  "20-39",
                                                  "40-59",
                                                  "60-79",
                                                  "80-99", 
                                                  "100+"))) %>% 
  left_join(a_lsoa_to_stp, by = c("Der_Postcode_LSOA_2011_Code" = "lsoa11cd")) %>% 
  left_join(a_stp_to_nhsr %>% 
              select(stp20cd, nhser20cd, nhser20nm) %>% 
              distinct(), 
            by = c("stp20cd"))

e_hip_ip_demo_index_unplanned <-
  a.ip_hip_unplanned %>% 
  #Ethnicity
  mutate(
    Ethnic_Group = case_when(Ethnic_Group %in% c("99", "NA") ~ Ethnic_Group,
                             TRUE ~ substr(Ethnic_Group,1,1)
    )) %>% # Shorten Ethnic group to first character to clean data (except when coded as 99 or NA)
  mutate(
    Ethnic_Group = case_when(is.na(Ethnic_Group) ~ "99",
                             TRUE ~ Ethnic_Group)) %>% # Replace NA's with "NA" to left_join to "Not Known"
  left_join(c_ethnicity_lookup, by = c("Ethnic_Group" = "Ethnic_Category")) %>% 
  # Deprivation 
  left_join(select(c_lsoa_imd, LSOA_Code, IMD_Score, IMD_Rank, IMD_Decile), by = c("Der_Postcode_LSOA_2011_Code" = "LSOA_Code")) %>% 
  mutate(IMD_Quintile = case_when(IMD_Decile %in% c(1,2) ~ 1,
                                  IMD_Decile %in% c(3,4) ~ 2,
                                  IMD_Decile %in% c(5,6) ~ 3,
                                  IMD_Decile %in% c(7,8) ~ 4,
                                  IMD_Decile %in% c(9,10) ~ 5,
                                  TRUE ~ 99)) %>% 
  # Activity count by demographic and region
  mutate(Admission_Date = as.Date(Admission_Date)) %>% 
  group_by(Admission_Date, nhser20nm, Age_range, Sex, Ethnic_Category_Desc, Ethnicity_broad, IMD_Quintile) %>% 
  summarise(Admission_count = n_distinct(APCS_Ident)) %>% 
  ungroup()

e_hip_ip_unplanned_week_number <-
  e_hip_ip_demo_index_unplanned %>% 
  mutate(Admission_week = floor_date(Admission_Date, unit = "week")) %>% 
  group_by(Admission_week) %>% 
  summarise(Admissions = sum(Admission_count)) %>% 
  mutate(year = year(Admission_week),
         #month = months(Admission_Date), 
         week_number = isoweek(Admission_week)) %>% 
  group_by(year, week_number) %>% 
  #summarise(weekly_admissions = sum(Admission_count)) %>% 
  ungroup()

# Annual trends
e_hip_ip_week_number_wide <-
  e_hip_ip_unplanned_week_number %>% 
  filter(year == 2020) %>% 
  rename(n_2020 = Admissions) %>% 
  select(-year, -Admission_week) %>% 
  left_join(e_hip_ip_unplanned_week_number %>% 
              filter(year == 2018) %>% 
              select(-year, -Admission_week) %>% 
              rename(n_2018 = Admissions), by = c("week_number"), keep = FALSE) %>% 
  left_join(e_hip_ip_unplanned_week_number %>% 
              filter(year == 2019) %>% 
              select(-year, -Admission_week) %>% 
              rename(n_2019 = Admissions), by = c("week_number"), keep = FALSE) %>% 
  group_by(week_number) %>% 
  mutate(Avg1819 =  round(mean(c(n_2018, n_2019), na.rm=T))) %>% 
  ungroup() %>% 
  mutate(def = Avg1819 - n_2020)

e_hip_total_admission_def_unplanned <-
  e_hip_ip_week_number_wide %>% 
  select(def) %>% 
  sum(na.rm = TRUE) 

# Sub-group proportions
e_get_admission_deficit_hip_unplanned <- function(demographic_var) {
  
  e_hip_ip_demo_index_unplanned <- 
    e_hip_ip_demo_index_unplanned %>%
    mutate(demographic_var = {{demographic_var}})
  
  sub_data <-
    e_hip_ip_demo_index_unplanned %>% 
    mutate(year = year(Admission_Date),
           month = months(Admission_Date), 
           week_number = isoweek(Admission_Date)) %>% 
    group_by(year, week_number, demographic_var) %>% 
    summarise(weekly_admissions = sum(Admission_count)) %>% 
    ungroup()
  
  sub_data_wide <-
    sub_data %>% 
    filter(year == 2020) %>% 
    rename(n_2020 = weekly_admissions) %>% 
    select(-year) %>% 
    left_join(sub_data %>% 
                filter(year == 2018) %>% 
                select(-year) %>% 
                rename(n_2018 = weekly_admissions), by = c("week_number", 'demographic_var'), keep = FALSE) %>% 
    left_join(sub_data %>% 
                filter(year == 2019) %>% 
                select(-year) %>% 
                rename(n_2019 = weekly_admissions), by = c("week_number", 'demographic_var'), keep = FALSE) %>% 
    group_by(week_number, demographic_var) %>% 
    mutate(Avg1819 =  round(mean(c(n_2018, n_2019), na.rm=T))) %>% 
    mutate(def = Avg1819 - n_2020) %>% 
    mutate(PC_def = case_when(
      week_number < 13 ~ 0,
      week_number >= 13 ~ Avg1819 - n_2020)) %>%  # Additional variable to start deficit from lockdown start
    ungroup() 
  
  sub_data_wide
}

# Cumulative deficit by region 
e_hip_unplanned_def_region <-
e_get_admission_deficit_hip_unplanned(nhser20nm) %>% 
  group_by(demographic_var) %>% 
  mutate(cum_def = cumsum(PC_def)) %>% 
  pivot_longer(cols = c(-week_number, -demographic_var),
               names_to = "trend", 
               values_to = "count") %>% 
  filter(trend == "cum_def") 

e_hip_ip_unplanned_def_region <-
e_get_admission_deficit_hip_unplanned(nhser20nm) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var) %>% 
  mutate(LCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100,2)) 

e_hip_ip_unplanned_def_deprivation <-
  e_get_admission_deficit_hip_unplanned(IMD_Quintile) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>% 
  group_by(demographic_var) %>% 
  mutate(LCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def, n = sum_Avg1819,0.95)$conf.int[2]*100,2)) 

e_hip_ip_unplanned_def_ethnicity <-
  e_get_admission_deficit_hip_unplanned(Ethnic_Category_Desc) %>% 
  group_by(demographic_var) %>% 
  summarise(sum_def = sum(PC_def, na.rm = TRUE),
            sum_Avg1819 = sum(Avg1819, na.rm = TRUE)) %>% 
  mutate(def_prop = round(sum_def/sum_Avg1819*100,2)) %>% 
  arrange(desc(def_prop)) %>%  
  mutate(sum_def_adj = case_when(sum_def >= 0 ~ sum_def,
                                 sum_def < 0 ~ -sum_def
                                 )) %>% # Switch out the negative def to positive to calcuate CI
  group_by(demographic_var) %>% 
  mutate(LCI = round(add4ci(x = sum_def_adj, n = sum_Avg1819,0.95)$conf.int[1]*100,2),
         UCI = round(add4ci(x = sum_def_adj, n = sum_Avg1819,0.95)$conf.int[2]*100,2)) %>% 
  mutate(LCI = case_when(sum_def >= 0 ~ LCI, sum_def < 0 ~ -LCI),
         UCI = case_when(sum_def >= 0 ~ UCI, sum_def < 0 ~ -UCI) 
         ) 


## F. QALY effects ####
# Admission deficit

# Age and gender profile 
f_hip_ip_Age_Gender_profile <-
a.ip_hip %>% 
  mutate(Admission_Date = as.Date(Admission_Date)) %>% 
  mutate(Period = case_when(
    (Admission_Date >= as.Date("2019-04-01") & Admission_Date <= as.Date("2019-12-31")) ~ "Apr_Dec_19",
    (Admission_Date >= as.Date("2020-04-01") & Admission_Date <= as.Date("2020-12-31")) ~ "Apr_Dec_20",
    TRUE ~ "Other")) %>% 
  filter(Period != "Other") %>% # Exclude not in periods of interest
  filter(Sex %in% c("1", "2")) %>% # Exclude 'Other' genders
  group_by(Sex, Age_at_Start_of_Episode_SUS, Period) %>% 
  summarise(Admission_count = n_distinct(APCS_Ident)) %>% 
  arrange(Age_at_Start_of_Episode_SUS, Period)

#write_csv(f_hip_ip_Age_Gender_profile, "hip_ip_Age_Gender_profile.csv")

f_life_expectancy <-
read_csv("life_expectancy.csv")

f_life_expectancy_discounted <-
  read_csv("life_expectancy_discounted.csv") %>% 
  clean_names()

# Pull in LE and LE discounted
f_hip_ip_Age_Gender_profile_LE <-
  f_hip_ip_Age_Gender_profile %>% 
  left_join(f_life_expectancy %>%
              filter(Gender == "Male") %>% 
              select(Age, ex), 
            by = c("Age_at_Start_of_Episode_SUS" = "Age"), keep = FALSE) %>% 
  rename(male_ex = ex) %>% 
  left_join(f_life_expectancy %>%
              filter(Gender == "Female") %>% 
              select(Age, ex), 
            by = c("Age_at_Start_of_Episode_SUS" = "Age"), keep = FALSE) %>% 
  rename(female_ex = ex) %>% 
  mutate(LE = case_when(Sex == 1 ~ male_ex,
                                     Sex == 2 ~ female_ex)) %>% 
  select(-male_ex, -female_ex) %>% 
  mutate(LE_round = round(LE,0)) %>% 
  left_join(f_life_expectancy_discounted %>% select(1,2), by = c("LE_round" = "le"))
  
# 
f_waiting_times <-
e_hip_ip_week_number_wide %>% 
  mutate(PC_def = case_when(
    week_number < 12 ~ 0,
    week_number >= 12 ~ Avg1819 - n_2020)) %>%  # Additional variable to start deficit from lockdown start
  mutate(cum_def = cumsum(PC_def)) %>% 
  mutate(weeks_left = 52 - week_number + 0.5) %>% 
  mutate(add_weeks_waiting = PC_def*weeks_left) %>% 
  mutate(cum_add_weeks_waiting = cumsum(add_weeks_waiting))

# Literature derived estimate of effect of 1 week delay in hip procedure on QALYs
f_qaly_loss_week <- 0.062/100 

#Total estimated QALY lost in population due to covid
f_total_qaly_loss <-
  sum(f_waiting_times$cum_add_weeks_waiting) * f_qaly_loss_week 

# Cost estimate
# 20,000 per QALY 
# 1 year
f_total_qaly_loss_value <- 
  f_total_qaly_loss *20000

# Life time
f_total_qaly_loss_value_lifetime <-
  f_total_qaly_loss_value * 
  (sum(f_hip_ip_Age_Gender_profile_LE$Admission_count) + 
     sum(f_hip_ip_Age_Gender_profile_LE$le_disc)) / 
  sum(f_hip_ip_Age_Gender_profile_LE$le_disc)


# Disutility waiting 
f_disutility_total <-
  sum(f_waiting_times$cum_add_weeks_waiting) * 19 

## G. London-style recovery model ####
# Regin specific data frames 

london_type_recovery <- function(.data) {
  df <- .data %>%
    pivot_longer(c(n_2020, Avg1819)) %>%
    group_nest(name) %>%
    mutate(fit = map(data, ~loess(value ~ week_number, .x, span = 0.2) %>%
                         predict(newdata = .x))) %>%
    mutate(across(data, map, select, week_number)) %>%
    unnest(c(data, fit)) %>%
    mutate(across(fit, pmax, 0)) %>%
    pivot_wider(names_from = name, values_from = fit) %>%
    mutate(london_type_recovery = case_when(
      week_number < 13 ~ n_2020,
      week_number < 24 ~ 0,
      week_number < 38 ~ sum(Avg1819 * as.numeric(week_number == 37)) / (37 - 24) * (week_number - 24),
      TRUE             ~ Avg1819
    )) %>%
    pivot_longer(-week_number)
  
  class(df) <- c("london_type_recovery", class(df))
  
  df
}

# Regional data frames
LTR_Midlands <-london_type_recovery(e_get_admission_deficit(nhser20nm) %>% 
                                      filter(demographic_var == "Midlands")) %>% 
  mutate(ID = "Midlands")

LTR_EastOfEngland <-london_type_recovery(e_get_admission_deficit(nhser20nm) %>% 
                                      filter(demographic_var == "East of England")) %>% 
  mutate(ID = "East Of England")

LTR_London <-london_type_recovery(e_get_admission_deficit(nhser20nm) %>% 
                                           filter(demographic_var == "London")) %>% 
  mutate(ID = "London")

LTR_NorthEast_Yorkshire <-london_type_recovery(e_get_admission_deficit(nhser20nm) %>% 
                                           filter(demographic_var == "North East and Yorkshire")) %>% 
  mutate(ID = "North East and Yorkshire")

LTR_NorthWest <-london_type_recovery(e_get_admission_deficit(nhser20nm) %>% 
                                           filter(demographic_var == "North West")) %>% 
  mutate(ID = "North West")

LTR_SouthEast <-london_type_recovery(e_get_admission_deficit(nhser20nm) %>% 
                                           filter(demographic_var == "South East")) %>% 
  mutate(ID = "South East")

LTR_SouthWest  <-london_type_recovery(e_get_admission_deficit(nhser20nm) %>% 
                                           filter(demographic_var == "South West")) %>% 
  mutate(ID = "South West")

LTR_union <-
LTR_Midlands %>% 
  union(LTR_London) %>% 
  union(LTR_NorthEast_Yorkshire) %>% 
  union(LTR_NorthWest) %>% 
  union(LTR_SouthEast) %>% 
  union(LTR_SouthWest) %>% 
  union(LTR_EastOfEngland) %>% 
  mutate(name = case_when(
    name == "Avg1819" ~ "2018-19",
    name == "london_type_recovery" ~ "Modelled",
    name == "n_2020" ~ "2020"
  ))


# Regional difference between 2020 trend and London-modelled trend
LTR_union_diff_london <-
LTR_union %>% 
  filter(name != "2018-19",
         ID != "London") %>% 
  pivot_wider(id_cols = c(week_number, ID),
              names_from = name,
              values_from = value) %>% 
  clean_names() %>% 
  mutate(diff_london = modelled - x2020) %>% 
  group_by(id) %>% 
  mutate(diff_london_cum = cumsum(diff_london)) 



```

# Executive summary 

<br>
National deficit of around 58k admissions for hip replacements between the onset of national lockdown (26th March 2020) and the end of 2020. Proportionally affecting the North West and South West NHSE regions most, where inpatient admissions during 2020 were over 50% less than would be expected. Lowest levels of disruption in London (approx. 30%) perhaps linked to differing approach to recovery where clinical hubs were set up to provide speciality-specific care.

<br>

**Admission deficit in planned care**: 

-	Elderly groups more disrupted (approx. 50% less admissions). Disruption reducing with age, and presumably linked to reducing risk perceptions.
-	Admission deficit/disruption proportionally highest in white British and other white ethnic sub-groups. 
-	Small but significant deprivation gradient seen in disruption to care, highest in most deprived, lowest in most affluent. This relationship is not consistent across regions; extenuated in London (30% care disrupted in IMD quintile 1, approx. 17% disrupted in IMD quintile 5), while the South East and South West display inverse deprivation relationships.

<br>

**Admission deficit in unplanned care**:

-	National deficit closely matches lockdown timeline
-	Unclear how to interpret a reduction in ED attendances  indicative of more effective planned care or reduced demand due to decreased mobility/activity. 
-	Highest proportional deficit in London, in comparison to lowest in London for planned admissions. 
-	Deprivation gradient seen in unplanned and planned admission disruption.

<br>
**COVID-19 impact on quality of life**

<br>
As of January 2021, the increased waiting for hip replacements resulted in a loss of approx 895 QALYs. Converting this figure to a monetary value, at 20k per QALY and discounting by 3.5% per year, values the total lifetime loss at 223 million. In addition, the disutility of increased waiting can be estimated at 27 million.



# Background 

The COVID-19 pandemic has had hugely disruptive impacts on the provision and delivery of emergency and elective care. Such effects have been felt differently across the country and have been moderated by disease progression and changing lockdown measures. Underlying health and social inequities across the population have been exposed. Pre-existing disease rates, barriers to healthcare and economic instability have aligned to present disproportionate threats to health for certain demographics and geographical regions.

## Hip replacements
GP referrals for orthopaedic elective care have seen one of the largest reductions during the pandemic adding to the already troubling waiting times for consultant led referrals [1]. Increased waiting times may lead to  increases in emergency admissions for untreated hip pain and mobility issues leading to trips, falls and acute presentations at A&E .  Emergency care systems may have increased thresholds by which patients are encouraged to attend; instead, patients were advised to seek virtual guidance or provide self-care where appropriate . Redirection of emergency care may have delayed attendance to A&E to the point of crisis, increasing the severity of such attendances, calling for more invasive treatment when seen and increasing costly disability during waiting periods [2]. The shift to virtual management of chronic hip pain and mobility issues has been vast but quality and uptake inevitably vary by provider and patient population, respectively. Virtual care brings potential for widening inequities when we consider differing economic accessibility, IT knowledge and healthcare literacy across age ranges and cultural groups. 

# Methods

## Project scope
The strategy unit proposes to quantify the changes during the COVD-19 pandemic in hip replacement activity, including pre- and post-operative care, according to selected demographic, socio-economic and geographic population groups.


## Data collection

### Inpatient care
Inpatient care was sub-divided into either 'elective' or 'unplanned' admissions based on the spell admission method. OPCS procedure codes were used to identify admissions with any of the following procedures:

Primary replacements:

-	Total Hip replacement
-	Total Prosthetic Replacement Head Femur 
-	Hybrid Prosthetic Hip 
-	Other Hip Replacement Primary
-	Resurfacing Arthroplasty Joint Primary

Revisions:

-	Total Hip Replacement Revision
-	Total Prosthetic Replacement Head Femur Revision
-	Hybrid Prosthetic Hip Replacement Revision
-	Other Hip Replacement Revision
-	Resurfacing Arthroplasty Joint Revision


### Outpatient appointments
Appointments including pre- or post-operative reviews relating to the above procedures and/or including any of the following diagnosis codes:

-	S700: Contusion of hip
-	S730: Dislocation of hip
-	S731: Sprain and strain of hip
-	S760: Injury of muscle and tendon of hip
-	S770: Crushing injury of hip
-	S772: Crushing injury of hip with thigh
-	S797: Multiple injuries of hip and thigh
-	S798: Other specified injuries of hip and thigh
-	S799: Unspecified injury of hip and thigh

### Emergency department attendances 
Unplanned attendances at emergency care settings for the following injuries/complaints were identified using SNOMED-CT codes:

-	Femur fracture (shaft and NOF)
-	Hip fractures (pubis, acetabulum, ilium, ischium)
-	Hip sprains 
-	Hip dislocation
-	Open wounds to the hip (lacerations, abrasions and open fractures)
-	Hip pain (these are findings as opposed to disorders)


## Analysis plan

Areas of interest include 

* What were the cohort characteristics of the patients who were prioritised (and inversely, those postponed) at the peak of the lockdown? 
  * Does this appear to be clinically driven?
* What were the cohort characteristics of the patients seeking emergency care? 
* How did the pandemic impact activity by care setting, disease type, demographic and geography? 
* What were the effects of delayed hip procedures on quality of life?


# COVID-19 impact on activity 
## National time series 
Fig. 1.1
``` {r Interactive dyGraph time series, warning = FALSE, message = FALSE}
dygraph(xts(x = b_hip_ip_daily$n_admissions, 
            order.by = b_hip_ip_daily$Admission_Date),
            main = "<b>Elective inpatient hip replacement admissions<b> <br> 
        <small> <i>Daily count, England 2018-20<i></small>",
        xlab = "Date", ylab = "Admissions"
        ) %>%
  dyOptions(labelsUTC = TRUE, 
            fillGraph=TRUE,  
            drawGrid = FALSE, 
            colors="#f9bf07") %>%
  dyRangeSelector() %>%
  dyCrosshair(direction = "both") %>%
  dyHighlight(highlightCircleSize = 10, 
              highlightSeriesBackgroundAlpha = 1, 
              hideOnMouseOut = FALSE)

``` 
<br>
The effect of COVID-19 on inpatient elective hip replacements is clear. The daily count of admissions fluctuated by weekday and weekend and there were regular reductions around Christmas and New Year Eve. A large drop is seen with the implementation of a national lockdown on 26th March 2020. Admission rates recovered and rose until September 30th 2020, after which daily admissions dropped gradually during the second national lockdown and the holiday period.  
<br>
<br>
<br>
Fig. 1.2
``` {r Hip activity by setting, warning = FALSE, message = FALSE}
b_hip_all_daily %>% 
  filter(Date > "2018-01-01" & 
           Date < "2020-12-31") %>% 
  mutate(week_day = weekdays(Date)) %>% 
  mutate(weekend_flag = case_when(substr(week_day,1,1) == "S" ~ "Weekend",
                                  TRUE ~ "Week day")) %>% 
  ggplot(aes(x = Date, y = Count)) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = 0, ymax = 600, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = 0, ymax = 600, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  annotate("text", x = as.Date("2020-05-13"), y = 500, label = "National lockdown",size = 3, angle = 90) +
  annotate("text", x = as.Date("2020-11-24"), y = 500, label = "3 Tier lockdown",size = 3, angle = 90) +
  #geom_point(aes(colour = weekend_flag), size = 1, alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.2) +
  facet_wrap(~Activity) +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73")) +
  scale_color_SU(palette = "main") +
  labs(title = "Planned hip replacement activity was halted by the pandemic",
       subtitle = "Hip replacement procedures and related activity: National, planned and unplanned care",
       x = "Date",
       y = "Activity")

```
<br>
COVID-19 impacted care settings differently; planned inpatient and outpatient care was canceled in line with national lockdown so trends follow similar patterns, while emergency department attendances saw less of a reduction but patients care seeking behaviour were clearly impacted to an extent. 
<br>
<br>
<br>

## Cohort characteristics 

### Age 
Fig. 2.1
``` {r Proporiton of hip procedures, by age range, warning = FALSE, message = FALSE, echo = FALSE}
ggplotly(
  c_ip_hip_age_vis +
    annotate("text", x = as.Date("2020-05-13"), y = 95, label = "National lockdown", size = 3, angle = 90) +
    annotate("text", x = as.Date("2020-11-24"), y = 85, label = "3 Tier lockdown", size = 3, angle = 90),
  tooltip = c("Admission_Date","prop")) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = paste0('<b>Daily proportion of hip procedures performed by age range<b>',
                                    '<br>',
                                    '<sup>',
                                    '<i>National, elective inpatient episodes<i>',
                                    '</sup>'))) 


```

Fig. 2.2.1 & 2.2.2
``` {r interactive Age Stacked bar, warning = FALSE, message = FALSE, echo = FALSE }
c_ip_hip_stack_and_line

```

### Deprivation 
Fig. 2.3
``` {r Proporiton of hip procedures, by deprivation, warning = FALSE, message = FALSE}
ggplotly(
  c.ip_hip_imd_vis,
  tooltip = c("Admission_week","prop")) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = paste0(
           '<b>During the initial lockown, hip care was focused on the very affluent and very deprived<b>',
           '<br>',
           '<sup>',
           '<i>Weekly proportion of hip procedures performed by IMD deprivation quintile: National, elective inpatient episodes<i>',
           '</sup>'))) 

```

Fig 2.3.2 - Steam graph 
```{r r Proporiton of hip procedures (steam graph), by deprivation, warning = FALSE, message = FALSE}
streamgraph(c_ip_hip_imd_wk %>% 
              filter(Admission_week > "2020-01-01" & 
                       Admission_week < "2020-12-31") %>% 
              ungroup(), 
            key="IMD_Quintile",
            value="sum",  
            date="Admission_week" #, height="300px", width="1000px"
            )
```


Fig. 2.4.1 & 2.4.2
``` {r IMD Stacked bar, warning = FALSE, message = FALSE}
#c_hip_ip_stacked_bar_and_line

subplot(
  # Stacked bar 
  ggplotly(c_ip_hip_imd_stack_vis +
             annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), y = 0, yend = 100, lty = 2) +
             annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), y = 0, yend = 100, lty = 2) +
             annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), y = 0, yend = 100, lty = 2) +
             annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), y = 0, yend = 100, lty = 2),
           tooltip = c("Admission_week","Proportion")
  ),
  
  # Line chart
  style(
    ggplotly(c_ip_hip_admission_volume_vis +
               annotate("segment", x = as.Date("2021-01-04"), xend = as.Date("2021-01-04"), y = 0, yend = 2800, lty = 2) +
               annotate("segment", x = as.Date("2020-03-23"), xend = as.Date("2020-03-23"), y = 0, yend = 2800, lty = 2) +
               annotate("segment", x = as.Date("2020-07-04"), xend = as.Date("2020-07-04"), y = 0, yend = 2800, lty = 2) +
               annotate("segment", x = as.Date("2020-10-14"), xend = as.Date("2020-10-14"), y = 0, yend = 2800, lty = 2),
             tooltip = c("Admission_week", "Admission_count")),
    showlegend = FALSE
  ),
  # Organise plots
  nrows = 2, heights = c(0.8,0.2), shareY = TRUE, shareX = TRUE 
) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = 
                        paste0(
                          '<b>More affluent people received a greater than average proportion of hip care during lockdown<b>',
                          '<br>',
                          '<sup>',
                          '<i>Proportion and count of elective hip procedures by deprivation quintile: National, weekly proportion<i>',
                          '</sup>'))) 

```

Fig. 2.5
``` {r IMD x Age range, warning = FALSE, message = FALSE, echo = FALSE}
c_ip_hip_imd_age_wk_vis +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1))
```

### Ethnicity
Fig. 2.7
``` {r Proporiton of hip procedures, by broad ethnicity, warning = FALSE, message = FALSE, echo = FALSE}
c_ip_hip_ethnicity_broad_vis %>% 
  ggplotly(tooltip = c("Ethnicity", "Proportion")) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = paste0(
           '<b>Weekly proportion of hip procedures performed by broad ethnicity<b>',
           '<br>',
           '<sup>',
           '<i>National, weekly average<i>',
           '</sup>')))
```

Fig. 2.8.1 & 2.8.2
``` {r stacked bar - broad ethnicity, warning = FALSE, message = FALSE, echo = FALSE}
c_hip_ip_stacked_bar_and_line_ethnicity
```

Fig. 2.9
``` {r Proporiton of hip procedures, by full ethnicity, warning = FALSE, message = FALSE, echo = FALSE}
c_ip_hip_ethnicity_wk_vis %>% 
  ggplotly(tooltip = c("Admission_week","Proportion", "Ethnicity")) %>% 
  layout(legend = list(y = 0.5),
         title = list(text = paste0(
           '<b>Weekly proportion of hip procedures performed by ethnicity<b>',
           '<br>',
           '<sup>',
           '<i>National, elective inpatient episodes<i>',
           '</sup>'))) 
```

## Deficit in planned admission 

### National and regional
Fig. 3.1 
``` {r Hip 2020 vs 2018-19 avg , warning = FALSE, message = FALSE}
e_hip_ip_annual_trends + e_hip_ip_admission_def_vis + 
  plot_annotation(title = "Yearly patterns in hip replacement admissions",
                        subtitle = "National, Weekly elective inpatient procedures, 2018-20"
                        )
``` 
<br>
By calculating the difference between the 2020 trend and the 2018-19 average weekly number of admissions we can present the admission deficit. So far, there have been `r e_hip_ip_total_admission_def` admissions less than we would expect for elective inpatient hip replacement admissions. The backlog of admissions is likely to last until after the deficit is zero.
<br>
<br>
<br>
Fig. 3.2
``` {r Hip 2020 vs 2018-19 by region, warning = FALSE, message = FALSE}
facet_plot_function(e_get_admission_deficit(nhser20nm)) +
  theme(legend.position = c(0.85, 0.2)) + 
  labs(x = "Week number", y = "Admissions", colour = "",
       title = "Admission deficit is determied by pre-pandemic admission volume and speed of 'recovery'",
       subtitle = "Elective inpatient hip admissions in 2020 compared to the average from 2018-19: Admissions by NHSE region")
```

```{r Hip 2020 vs 2018-19 by region - table, warning = FALSE, message = FALSE}
create_dt(
  e_get_admission_deficit(nhser20nm)
)
```

```{r create_dt function, warning = FALSE, message = FALSE}
create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                               lengthMenu = list(c(7, 10,50,-1),
                                                 c(7, 10,50,"All"))))
}
```


<br>
The initial effect of covid-19 and national lockdown brought elective inpatient admissions for hip replacements down to near zero in all areas. As such, high admission deficits are seen where the admission volume was high pre-pandemic and volumes did not return to the levels seen in previous years through the 'Recovery' period; specifically Midlands and  North East and Yorkshire.
<br>
<br>
<br>
Fig. 3.3
``` {r Admission deficit by region, warning = FALSE, message = FALSE}
e_weekly_cum_def_by_region %>%
  filter(week_number < 53) %>% 
  mutate(label = if_else(week_number == max(week_number), as.character(demographic_var), NA_character_)) %>% 
  ggplot(aes(x = week_number, y = count, colour = demographic_var)) +
  annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
  annotate("text", x = 20, y = -100, label = str_wrap("Lockdown", 10) ,size = 3) +
  annotate("text", x = 34, y = -100, label = str_wrap("Recovery", 10) ,size = 3) +
  annotate("text", x = 48, y = -100, label = str_wrap("Lockdown", 10) ,size = 3) +
  geom_smooth(method = 'loess', span = 0.2) +
  geom_label_repel(aes(label = label), nudge_x = 1, size = 3.5,  na.rm = TRUE) +
  scale_color_SU(palette = "main") +
  theme(legend.position = "none") +
  labs(title = "The Midlands region has the highest absolute admission deficit for hip procedures", 
       subtitle = "Cumulative admission deficit for hip replacement procedures due to Covid-19 pandemic",
       x = "Week number",
       y = "Admission deficit")

```
<br>
Another influential factor in determining the volume of a region's admission deficit is the magnitude of the second lockdown. Again the Midlands and North East and Yorkshire experienced particularly distruptive second lockdowns, as seen above.
<br>
<br>
<br>
Fig. 3.4
``` {r Proportion deficit by region - graph, warning = FALSE, message = FALSE}
e_hip_ip_total_deficit_region %>% 
  ggplot(aes(y = reorder(demographic_var, def_prop), x = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(xmin = LCI, xmax = UCI), width = 0.25, colour = "#636363") +
  coord_cartesian(xlim=c(20,60)) +
  labs(title = "Regional inequalities are clear between the East, South East and the rest",
       subtitle = "Admission deficit due to covid-19 pandemic, by NHSE region",
       y = "NHSE region", 
       x = "Admission deficit proportion", 
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19")

```
<br>
'Deficit proportion' is calculated below by dividing the region's admission deficit by the average annual number of admissions seen between 2018 and 2019, as a proxy for the number of admissions we would have expected to see in the absence of the pandemic.
<br>
The above chart demonstrates that while the most patients were delayed or cancelled in the Midlands and North East and Yorkshire, the largest proportional effect of covid-19 on elective hip replacement admissions was seen in the North West; where over half of the expected activity was disrupted. Conversely, London and the South East achieved the lowest admission deficit proportion with only 30-40% of the expected number of admissions delayed or cancelled.
<br>
<br>
<br>

### Demographic deficits

#### Age
<br>
Fig. 3.5
``` {r Admission deficit proportion by age range, warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_def_age_range %>%
  ggplot(aes(x = reorder(demographic_var, def_prop), y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  coord_flip() +
  labs(title = "Though initially prioritised for hip procedures, eldlerly age groups were most delayed or cancelled",
       subtitle = "Admission deficit due to covid-19 pandemic, by age range",
       x = "Age range", 
       y = "Admission deficit proportion", 
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19")
```
<br>
Though a spike in the proportion of admissions attributed to 80-99 year old's was seen at the onset of national lockdown, the overall deficit demonstrates high levels of disruption in this group with over half of the expected admission volume for this group either delayed or cancelled. This effect is seen to deminish as you move down the age ranges, where the least disruption is seen in age ranges considered to have the lowest risk of covid-19 morbidity and mortality.
<br>
<br>
<br>

#### Ethnicity 
<br>
Fig. 3.6
``` {r Admission deficit - ethnicity, warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_def_ethnicity %>%
  ggplot(aes(x = reorder(demographic_var, def_prop), y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  coord_flip() +
  labs(title = "White people were most disrupted by delayed or cancelled hip replacement care",
       subtitle = "Admission deficit due to covid-19 pandemic, by ethnicity",
       x = "Ethnicity", 
       y = "Admission deficit proportion",
       caption = "Note: Deficit as a proportion of total expected admissions seen in '18 and '19")

```
<br>
Ethnicities classified broadly as 'White' are the three most delayed or cancelled ethnic groups; 'British', 'Irish' and 'Any other white background'. A large effect was also seen in the 'Not Stated' group, suggesting around 35% of the expected activity for this group was disrupted. Mixed effects were seen elsewhere, making conclusions by broad ethnic groups difficult.
<br>
<br>
<br>
Fig. 3.7 
``` {r Proportional change in admissions by ethnicity, warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_def_ethnicity_prop_change %>%
  ggplot(aes(x = demographic_var)) +
  geom_segment(aes(x = demographic_var, xend = demographic_var, y = 0, yend = prop_change)) +
  geom_point(aes(y = prop_change), fill = "orange", colour = "black",  shape = 21, size = 4) +
  coord_flip() +
  labs(title = "The greatest proportional increases in hip admissions performed were in 'Unknown' ethnicities", 
       subtitle = "Chaning proportion of planned, inpatient hip procedures before and during the pandemic, by ethnicity",
       x = "Ethnicity", 
       y = "Percentage point difference",
       caption = "Note: Proportion from 2020 compared to 2018-19 average") 
```
<br>
The respective representation of ethnic groups was largely the same between 2018-19 and 2020 other than increases seen in 'Unknown' ethnicities and relative decreases in the British group. Changes seen here may be linked by data collection practices during high stress periods, in that a large proportion of the increase in unknown may be coming from reductions in British patients. 
<br>
<br>
<br>
Fig 3.8 
``` {r NHSE region x ethnicity, warning = FALSE, message = FALSE, echo = FALSE}
e_deficit_nhsr_ethnicity_prop_change %>% 
  ggplot(aes(x = demographic_var2)) +
  geom_segment(aes(x = demographic_var2, xend = demographic_var2, y = 0, yend = prop_change)) +
  geom_hline(aes(yintercept = 0)) +
  geom_point(aes(y = prop_change), fill = "orange", colour = "black",  shape = 21, size = 6) +
  facet_wrap(~demographic_var1) +
  coord_flip() +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Changes in White and Unknown ethnicity proportions are consistent across the country", 
       subtitle = "Changing proportion of planned, inpatient hip procedures from before to during the pandemic, by ethnicity and region",
       x = "Broad ethnicity", 
       y = "Percentage point difference",
       caption = "Note: Proportion from 2020 compared to 2018-19 average") 
```
<br>
Trends in the changing proportions of hip replacement procedures performed by ethnicity at the national level are echoed across the country. Reducing proportions of white hip replacement patients and increasing proportions of patients of 'unknown' ethnicity vary in terms of magnitude regionally but are directionally consistent. 
<br>
<br>
<br>

#### Deprivation

<br>
Fig. 3.9
``` {r Admission deficit by deprivation quintile, warning = FALSE, message = FALSE}
# Deprivation
e_hip_ip_def_deprivation %>% 
  ggplot(aes(x = reorder(demographic_var, -demographic_var), y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  geom_hline(yintercept = 48.5, linetype = "dashed") +
  coord_flip() +
  labs(title = "A clear deprivation gradient is present in disruption to hip replacement admissions",
       subtitle = "Admission deficit due to covid-19 pandemic, by IMD Quintile",
       x = "IMD Quintile", 
       y = "Admission deficit proportion", 
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19")
```
<br>
Hip replacement admissions for the most deprived 20% of the population were significantly more disrupted than each of the more affluent IMD quintiles.
<br>
<br>
<br>
Fig. 3.10
``` {r Change in prop between 2020 and 201819 - IMD, warning = FALSE, message = FALSE, echo = FALSE}
# Comparing proportion of hip procedures by deprivation between 2018/19 avg and 2020
e_hip_ip_def_deprivation_change %>%
  ggplot(aes(x = demographic_var)) +
  geom_segment(aes(x = demographic_var, xend = demographic_var, y = 0, yend = prop_change)) +
  geom_point(aes(y = prop_change), fill = "orange", colour = "black",  shape = 21, size = 6) +
  geom_hline(aes(yintercept = 0)) +
  labs(title = "The proportions of people who did recieve hip replacements changed to benefit the affluent", 
       subtitle = "Changing proportion of planned, inpatient hip procedures before and during the pandemic, by deprivation",
       x = "IMD Quintile", 
       y = "Percentage point difference",
       caption = "Note: Proportion from 2020 compared to 2018-19 average") 

```
<br>
A deprivation gradient exists when comparing the changing proportion of elective hip replacement procedures between the 2018-19 average and those admitted during 2020. The proportion is seen to increase in more affluent groups and decrease is the more deprived members of the population. Where initially we saw a small increase in the proportion of admissions attributed to IMD quintile 1 (the most deprived 20% of the population) during the first lockdown, we then saw a generally decreasing trend for this group through the 'recovery' and 'second lockdown' periods. 
<br>
<br>
<br>
Fig. 3.11
``` {r NHSE region x Deprivation, warning = FALSE, message = FALSE}
# NHSE region x Deprivation
e_hip_ip_nhserXdeprivation %>% 
  ggplot(aes(x = demographic_var2, y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  facet_grid(~str_wrap(demographic_var1,12), scales = "free") +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73")) +
  labs(title = "In London, the deprived were much more disrupted than the affluent",
       subtitle = "Admission deficit for hip replacements due to covid-19 pandemic, by age range",
       x = "IMD Quintile", 
       y = "Admission deficit proportion",
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19")

```
<br>
The patterns we see in admission deficit by deprivation quintile vary across the country. London demonstrates clear inequity with deprived people having their procedures delayed or cancelled much more than affluent people. It may also be the case that more deprived groups weren't demanding care as much as the affluent were, potentially due to lack of sick leave/pay or more hesitancy to attend hospital during the pandemic; similar gradients can be seen in the North East and Yorkshire and the North West regions respectively. Conversely it seems that the case of more affluent populations were most disrupted by the pandemic in the South East and South West respectively.  
<br>
<br>
<br>
Fig. 3.12
``` {r Deprivation x ethnicity, warning = FALSE, message = FALSE}
# Deprivation x Broad ethnicity
e_hip_ip_def_deprivationXEthnicity_broad %>% 
  ggplot(aes(x = demographic_var1, y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  facet_grid(~str_wrap(demographic_var2, 12), scales = "free") +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73")
        ) +
  labs(title = "Inequities in disruption by deprivation are most apparent in Asian patients",
       subtitle = "Admission deficit due to covid-19 pandemic, by ethnicity and deprivation",
       x = "IMD Quintile", 
       y = "Admission deficit proportion",
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19")

```
<br>
Deprivation gradients are present in Asian and White groups; to varying degrees, the more deprived members of these ethnic groups were more disrupted by the pandemic. 
<br>
<br>
<br>
Fig. 3.13
``` {r  Deprivation x ethnicity - Asian specific, warning = FALSE, message = FALSE}
e_hip_ip_def_deprivationXEthnicity_broad_Asain %>%   
  ggplot(aes(x = demographic_var1, y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  facet_grid(~str_wrap(demographic_var2,15)) +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73")) +
  labs(title = "Admission deficit due to covid-19 pandemic",
       subtitle = "Deficit as a proportion of total expected admissions seen in '18 and '19",
       x = "IMD Quintile", 
       y = "Admission deficit proportion")
```
<br>
Deprivation inequities in Asian populations seem to be driven by Pakistani and Indian subgroups. Where negative deficit proportions are seen, we can deduce excess admissions occurred; more admissions in this subgroup in 2020 than the 2018-19 average. This is the case with Bangladeshi patients however the magnitude of the excess is very small and therefore has little influence on the aggregated results for the Asian population.      
<br>
<br>
<br>
Fig. 3.14
``` {r Deprivation x Broad ethnicity - change in proportion, warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_def_deprivationXEthnicity_broad_change_prop %>%
  ggplot(aes(x = demographic_var1)) +
  geom_segment(aes(x = demographic_var1, xend = demographic_var1, y = 0, yend = prop_change)) +
  geom_hline(aes(yintercept = 0)) +
  geom_point(aes(y = prop_change), fill = "orange", colour = "black",  shape = 21, size = 6) +
  facet_wrap(~demographic_var2) +
  coord_flip() +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Trends in White and Unknown proportions are consistent across deprivation quintile ", 
       subtitle = "Changing proportion of planned, inpatient hip procedures from before to during the pandemic, by ethnicity and deprivation",
       x = "IMD Quintile", 
       y = "Percentage point difference",
       caption = "Note: Proportion from 2020 compared to 2018-19 average") 
```
<br>
Increases in proportion of those receiving hip replacements are seen in all ethnicities other than the White population. The magnitude of the increase is much higher in the 'Not known' category than other non-White groups. The deprivation profile of the 'Not known' ethnicity appears to inversely match that of the White population supporting the hypothesis that increases in 'Not known' are primarily from reductions in the White patients.
<br>
<br>
<br>
Fig. 3.15
``` {r Admission deficit - Deprivation x Age, warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_def_deprivation_age %>%
  ggplot(aes(x = demographic_var1, y = proportion)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  facet_grid(~str_wrap(demographic_var2, 12), scales = "free") +
  theme(strip.text = element_text(size = 10, face = "bold"),
        strip.background = element_rect(colour = "#686f73")
  ) +
  labs(title = "Inequalities according to deprivation are most clear in 20-39 year olds",
       subtitle = "Admission deficit due to covid-19 pandemic, by age range",
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19",
       x = "IMD Quintile", 
       y = "Admission deficit proportion")

```
<br>
Inequalities in hip replacement admission deficit appear to be most accentuated in the 20-39 year old age range. Here we can see the most deprived members of the population are disproportionately disrupted while there is less of a deprivation gradient in the 60+ population, where the most hip replacements are performed. 
<br>
<br>
<br>

## Deficit in unplanned admissions

### National and regional

Fig. 4.1
``` {r Unplanned hip replacement activity - time series, warning = FALSE, message = FALSE}
# Time series
e_hip_ip_demo_index_unplanned %>% 
  mutate(Admission_week = floor_date(Admission_Date, unit = "week")) %>% 
  group_by(Admission_week) %>% 
  summarise(Admissions = sum(Admission_count)) %>% 
  filter(Admission_week > "2018-01-01" &  Admission_week < "2020-12-31") %>%
  
  ggplot(aes(x = Admission_week, y = Admissions)) +
  annotate("rect", xmin = as.Date("2020-03-23"), xmax = as.Date("2020-07-04"), ymin = -Inf, ymax = Inf, 
           fill= "#fed976", alpha = 0.8) + # Lockdown 1
  annotate("rect", xmin = as.Date("2020-10-14"), xmax = as.Date("2021-01-04"), ymin = -Inf, ymax = Inf, 
           fill= "#fed976", alpha = 0.8) + # 3 tier system
  annotate("text", x = as.Date("2020-05-13"), y = 1000, label = "National lockdown",size = 3.5, angle = 90) +
  annotate("text", x = as.Date("2020-11-24"), y = 1000, label = "3 Tier lockdown",size = 3.5, angle = 90) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.2) +
  labs(title = "Unplanned hip replacements were largely unaffected by the pandemic",
       subtitle = "Weekly admissions for unplanned emergency hip replacement activity: National 2018-20",
       x = "Date",
       y = "Unplanned admissions")

```
<br>
When viewing the trend in emergency, unplanned admissions for hip replacements, we see regular winter spikes in December 2018 and 2019. We also see a small reduction from the normal activity level around the first national lockdown with  a gradual 'recovery'
<br>
<br>
<br>
Fig. 4.2
``` {r Unplanned hip replacement - annual trends,  warning = FALSE, message = FALSE}
e_hip_ip_week_number_wide %>%   
  ggplot(aes(x = week_number, y = n_2020)) +
  # Covid timeline markers
  annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
  annotate("text", x = 20, y = 1050, label = str_wrap("National lockdown", 10) ,size = 3.5) +
  annotate("text", x = 34, y = 1050, label = str_wrap("Recovery period", 10) ,size = 3.5) +
  annotate("text", x = 48, y = 1050, label = str_wrap("3 Tier lockdown", 10) ,size = 3.5) +
  ## Labels and arrows 
  annotate(geom = "curve", x = 20, y = 1000, xend = 21.5, yend = 925, 
           curvature = -0.3, arrow = arrow(length = unit(2, "mm"))) +
  annotate(geom = "curve", x = 19, y = 750, xend = 16, yend = 775, 
           curvature = -0.3, arrow = arrow(length = unit(2, "mm"))) +
  annotate("text", x = 18, y = 1000, label = str_wrap("2018-19 average",10), size = 3.5) +
  annotate("text", x = 20, y = 750, label = "2020", size = 3.5) +
  geom_smooth(method = "loess", span = 0.2, colour = "#f9bf07") +
  geom_smooth(aes(y = Avg1819), method = "loess", span = 0.2, colour = "#5881c1") +
  labs(x = "Week number", y = "Admissions") +
  labs(title = "Both lockdown periods generate unplanned hip replacement admission deficits",
       subtitle = "Annual trends in weekly admissions for unplanned emergency hip replacement activity: National 2018-20",
       x = "Week number",
       y = "Weekly admissions")
  
```
<br>
Outside of national lockdown periods, we see very similar admission rates for unplanned hip replacement procedures. Factors that may be influential here are a) reduced population mobility (similar to trends seen in RTA's?), b) changing ED thresholds for admitting patients who attend with hip injuries and/or c) reduced care seeking behaviour.
<br>
<br>
<br>
Fig. 4.3
``` {r Unplanned hip replacement - cumulative deficit, warning = FALSE, message = FALSE}
# Cumulative deficit
e_hip_ip_week_number_wide %>% 
  mutate(cum_def = cumsum(def)) %>% 
  ggplot(aes(x = week_number, y = cum_def)) +
  geom_smooth(method = "loess", span = 0.2, colour = "#f9bf07") +
  #geom_line() +
  # Covid timeline markers
  annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
  annotate("text", x = 20, y = 250, label = str_wrap("National lockdown", 10) ,size = 4) +
  annotate("text", x = 34, y = 250, label = str_wrap("Recovery period", 10) ,size = 4) +
  annotate("text", x = 48, y = 250, label = str_wrap("3 Tier lockdown", 10) ,size = 4) +
  labs(x = "Week number",
       y = "Cumulative deficit",
       title = "Reductions from the norm in unplanned hip replacements closely match the lockdown timeline", 
       subtitle = "Cumulative unplanned admission deficit for unplanned hip replacement, National",
       caption = "Deficit = difference between 2020 admission count and expected admissions seen in '18 and '19")
```
<br>
The cumulative deficit trend above shows that while admissions rates similar to the previous years returned during the 'recovery' period, the deficit will remain until rates surpass the comparative trend. 
<br>
<br>
<br>
Fig. 4.4
``` {r Unplanned hip replacement - cumulative deficit by region, warning = FALSE, message = FALSE}
e_hip_unplanned_def_region %>%
  filter(week_number < 53) %>% 
  mutate(label = if_else(week_number == max(week_number), as.character(demographic_var), NA_character_)) %>% 
  ggplot(aes(x = week_number, y = count, colour = demographic_var)) +
  annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
  annotate("text", x = 20, y = -100, label = str_wrap("Lockdown", 10) ,size = 3) +
  annotate("text", x = 34, y = -100, label = str_wrap("Recovery", 10) ,size = 3) +
  annotate("text", x = 48, y = -100, label = str_wrap("Lockdown", 10) ,size = 3) +
  geom_hline(yintercept = 0) +
  geom_smooth(method = 'loess', span = 0.2) +
  geom_label_repel(aes(label = label), nudge_x = 3, size = 3.5,  na.rm = TRUE) +
  scale_color_SU(palette = "main") +
  theme(legend.position = "none") +
  labs(title = "London and the South West have consistently had the largest unplanned admission deficits", 
       subtitle = "Admission deficit for unplanned hip replacement procedures due to Covid-19 pandemic",
       x = "Week number",
       y = "Admission deficit")

```
<br>

<br>
<br>
<br>
Fig. 4.4.2 (alternative plot)
``` {r Unplanned hip replacement - cumulative deficit by region - alternative, warning = FALSE, message = FALSE, echo = FALSE}
  e_hip_unplanned_def_region %>%
    filter(week_number < 53) %>% 
    
    ggplot(aes(x = week_number, y = count)) +
    annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
    annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
    annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
    annotate("text", x = 20, y = -100, label = str_wrap("Lockdown", 10) ,size = 3) +
    annotate("text", x = 34, y = -100, label = str_wrap("Recovery", 10) ,size = 3) +
    annotate("text", x = 48, y = -100, label = str_wrap("Lockdown", 10) ,size = 3) +
    geom_hline(yintercept = 0) +
    geom_smooth(method = 'loess', span = 0.2, colour = "#f9bf07") +
    facet_wrap(~demographic_var) +
    theme(legend.position = "none",
          strip.text = element_text(face = 'bold')) +
    labs(title = "London and the South West have consistently had the largest unplanned admission deficits", 
       subtitle = "Admission deficit for unplanned hip replacement procedures due to Covid-19 pandemic",
       x = "Week number",
       y = "Admission deficit")
```
<br>
...
<br>
<br>
<br>
Fig 4.5
``` {r Admission deficit proportion by region, warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_unplanned_def_region %>%
  ggplot(aes(y = reorder(demographic_var, def_prop), x = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(xmin = LCI, xmax = UCI), width = 0.25, colour = "#636363") +
  #coord_cartesian(xlim=c(20,60)) +
  labs(title = "Unplaned admission deficit due to covid-19 pandemic, by NHSE region",
       subtitle = "Deficit as a proportion of total expected admissions seen in '18 and '19",
       x = "Unplanned admission deficit proportion", 
       y = "NHSE region")
```
<br>
There are clear and significant differences in disruption to emergency hip replacements across the country; it is unclear whether differences here are due to region differences in maintaining  planned procedures rates throughout the pandemic or changing emergency department thresholds to admit patients.
<br>
<br>
<br>
Fig. 4.6
``` {r Admissin def scatter - region, warning = FALSE, message = FALSE}
# Region - planned vs unplanned admissions
e_hip_ip_total_deficit_region %>% 
  select(demographic_var, def_prop) %>% 
  rename(planned_def = def_prop) %>% 
  left_join(e_hip_ip_unplanned_def_region %>% 
              select(demographic_var, def_prop) %>% 
              rename(unplanned_def = def_prop), 
            by = c("demographic_var")) %>%
  ungroup() %>% 
  
  ggplot(aes(y = planned_def, x = unplanned_def, label = str_wrap(demographic_var, 12))) + 
  geom_hline(aes(yintercept = mean(planned_def)), linetype = "dashed", colour = "#636363") +
  annotate("text",x = 6.8, y = 46.5, label = "Planned Average",size = 3) +
  geom_vline(aes(xintercept = mean(unplanned_def)), linetype = "dashed", colour = "#636363") +
  annotate("text", x = 4.8, y = 40, label = "Unplanned Average", size = 3, angle = 90) +
  geom_point(fill = "#f9bf07", colour = "black", pch = 21, size = 8) +
  geom_text_repel(box.padding = 1) +
  labs(title = "Distruption of hip replacements is most pronouced in the South West NHSE region ",
       subtitle = "Admission deficit proportion by region and admission type",
       x = "Unplanned deficit proportion",
       y = "Planned deficit proportion ")
 
```
<br>
By comparing each region's planned and unplanned admission deficit proportion we see two main outliers: London with a high unplanned deficit alongside a low planned deficit, while the South East has low planned and unplanned deficits. Admission rates in the South East have remained the most consistent with previous years in terms of planned and unplanned hip replacements. It is unclear from this analysis how the two rates interact and co-depend. 
<br>
<br>

### Demographic deficits

##### Deprivation

<br>
Fig. 4.7 
``` {r Admission deficit proportion by deprivation, warning = FALSE, message = FALSE}
e_hip_ip_unplanned_def_deprivation %>%
  ggplot(aes(x = reorder(demographic_var, -demographic_var), y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25, colour = "#636363") +
  coord_flip() +
  labs(title = "The most affluent 40% had the smallest proportional reduction in unplanned hip replacements",
       subtitle = "Unplanned hip replacement admission deficit due to covid-19 pandemic, by deprivation",
       y = "Admission deficit proportion", 
       x = "IMD Quintile",
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19"
       )
```
<br>
We can see that the more affluent members of the population had the least disruptions to the unplanned admission rate. Attributing that trend to policy or behaviour is difficult. Were the more affluent groups more able to navigate the care system to demand care in spite of the pandemic? Were more deprived groups more apprehensive to seek care for fear of Covid-19?
<br>
<br>
Fig. 4.8
``` {r Admission def scatter - deprivation,  warning = FALSE, message = FALSE}
e_hip_ip_def_deprivation %>% 
  select(demographic_var, def_prop) %>% 
  rename(planned_def = def_prop) %>% 
  left_join(e_hip_ip_unplanned_def_deprivation %>% 
              select(demographic_var, def_prop) %>% 
              rename(unplanned_def = def_prop), 
            by = c("demographic_var")) %>%
  ungroup() %>% 
  
  ggplot(aes(y = planned_def, x = unplanned_def, label = str_wrap(demographic_var, 12))) + 
  geom_hline(aes(yintercept = mean(planned_def)), linetype = "dashed", colour = "#636363") +
  annotate("text",x = 3.5, y = 48.4, label = "Planned Average",size = 3.5) +
  geom_vline(aes(xintercept = mean(unplanned_def)), linetype = "dashed", colour = "#636363") +
  annotate("text", x = 4.1, y = 49.5, label = "Unplanned Average", size = 3.5, angle = 90) +
  geom_point(fill = "#f9bf07", colour = "black", pch = 21, size = 10) +
  geom_text_repel(box.padding = 1.5) +
  labs(title = "Inequities in hip replacement disruption are consistent across admission types",
       subtitle = "Admission deficit proportion by deprivation quintile and admission type",
       x = "Unplanned deficit proportion",
       y = "Planned deficit proportion ")
```
<br>
Viewing changes in planned admission deficits alongside that of unplanned care might identify reductions in one admission type and subsequent increases in the other; however when viewing the population by deprivation quintile, we see the more affluent people experienced the least disruption to planned and unplanned hip replacement care. 
<br>
<br>

##### Ethnicity

<br>
Fig. 4.9
``` {r Admission deficit proportion by ethnicity , warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_unplanned_def_ethnicity %>%
  ggplot(aes(x = reorder(demographic_var, def_prop), y = def_prop)) +
  geom_col(fill = "#f9bf07", colour = "grey") +
  geom_errorbar(aes(ymin = LCI, ymax = UCI), width = 0.25 , colour = "#636363") +
  coord_flip() +
  labs(title = "Trends in unplanned admission deficits do not match those of planned admissions",
       subtitle = "Unplanned hip replacement admission deficit due to covid-19 pandemic, by ethnicity",
       y = "Admission deficit proportion", 
       x = "Ethnicity",
       caption = "Deficit as a proportion of total expected admissions seen in '18 and '19")
```
<br>
Broad confidence intervals suggest drawing conclusions from ethnicity groups with small numbers may be ill-advised. We can say with confidence that the emergency hip care for Pakistani, Indian and Caribbean members of the population was disproportionately disrupted, in comparison to British and other white groups.
<br>
<br>
Fig. 4.10
``` {r Admission def scatter - Ethnicity, warning = FALSE, message = FALSE, echo = FALSE}
e_hip_ip_def_ethnicity %>% 
  mutate(conf = UCI - LCI) %>% 
  select(demographic_var, def_prop, conf) %>% 
  rename(planned_def = def_prop) %>% 
  left_join(e_hip_ip_unplanned_def_ethnicity %>% 
              mutate(conf = UCI - LCI) %>% 
              select(demographic_var, def_prop, conf) %>% 
              rename(unplanned_def = def_prop), 
            by = c("demographic_var")) %>%
  mutate(confidence = mean(conf.x, conf.y)) %>% 
  ungroup() %>%
  mutate(conf_tertile = 
           case_when(
             confidence < quantile(confidence, probs = 0.33) ~ "High",
             confidence > quantile(confidence, probs = 0.33) & confidence < quantile(confidence, probs = 0.66) ~ "Medium",
             confidence > quantile(confidence, probs = 0.66) ~ "Low"
           ))  %>% 
  mutate(conf_tertile = factor(conf_tertile, levels = c("Low", "Medium", "High"))) %>% 
  
  ggplot(aes(y = planned_def, x = unplanned_def, label = str_wrap(demographic_var, 12))) + 
  geom_hline(aes(yintercept = mean(planned_def)), linetype = "dashed", colour = "#636363") +
  annotate("text",x = 11, y = 26, label = "Planned Average",size = 3.5) +
  geom_vline(aes(xintercept = mean(unplanned_def)), linetype = "dashed", colour = "#636363") +
  annotate("text", x =5.1, y = 40, label = "Unplanned Average", size = 3.5, angle = 90) +
  geom_point(aes(alpha = conf_tertile), fill = "#f9bf07", colour = "black", pch = 21, size = 10) +
  geom_text_repel(box.padding = 1, size = 3) +
  labs(title = "Proportional disruption in hip replacement care was highest for Pakistani and Indian groups",
       subtitle = "Admission deficit proportion by ethnicity and admission type",
       x = "Unplanned deficit proportion",
       y = "Planned deficit proportion",
       alpha = "Confidence")
```
<br>
When comparing proportional admission deficit for planned and unplanned care, it is clear the Asian sub-groups are experiencing higher than average levels of disruption in planned and unplanned admissions for hip replacements and related care; the same could be said for white British populations.
<br>
<br>

# Quality of life effects 

As of January 2021, the increased waiting for hip replacements resulted in a loss of approx __*895*__ QALYs. Converting this figure to a monetary value, at __*20k*__ per QALY and discounting by __*3.5%*__ per year, values the total lifetime loss at __*223 million*__. In addition, the disutility of increased waiting can be estimated at __*27 million*__. 
<br>
<br>
These figures will continue to rise until the deficit is reduced, that is we recover all the activity lost due to covid-19. Details below:
<br>
<br>

- Literature derived [2] QALY impact of 1 week delay in hip replacement surgery: __*0.00062*__ 
- Total admission deficit: `r paste(e_hip_ip_total_admission_def, "admissions")`
- Cumulative additional weeks waiting for hip replacement: `r paste(f_waiting_times$cum_add_weeks_waiting[f_waiting_times$week_number == 52], "weeks")`
- QALYs lost by those waiting for hip procedures: `r paste(round(f_total_qaly_loss,1), "QALYS")`
- Monetary value of QALYs lost by those waiting (year 1): `r paste0("",round(f_total_qaly_loss_value,0))`
- Monetary value of lifetime lost of QALYs: `r paste0("",round(f_total_qaly_loss_value_lifetime,0))`
- Disutility of waiting for hip procedures: `r paste0("",round(f_disutility_total,0))`

<br>
<br>
<br>

# Adittional analysis

## London-style recovery modelling

<br>
**How would the national deficit in elective hip replacements vary if all regions had recovered like London?**

<br>
We modelled an alternative admission trend for each region to match the recovery profile seen in London; that is, near-zero admission rates throughout the first lockdown, and a steady recovery to pre-pandemic rates by week 38.

The model can be described as:

$$
y = \begin{cases}
    p1_x, & \text{if $x$ prior to lockdown}.\\
    0, & \text{if $x$ in lockdown}.\\
    m(x-X), & \text{if $x$ in recovery}.\\
    p2_x, & \text{otherwise}
  \end{cases}
$$
where:

  * $X$ is the last week in lockdown, 
  * $x$ is the week,  
  * $p1_x$ is the 2020 value
  * $p2_x$ is the 2018-19 value

<br>
```{r London-style recovery, warning = FALSE, message = FALSE}
LTR_union %>% 
  ggplot(aes(week_number, value, colour = name)) +
  annotate("segment", x = 13, xend = 13, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 27, xend = 27, y = -Inf, yend = Inf, lty = 2) +
  annotate("segment", x = 42, xend = 42, y = -Inf, yend = Inf, lty = 2) +
  annotate("text", x = 20, y = 790, label = str_wrap("Lockdown", 10) ,size = 2.5) +
  annotate("text", x = 34, y = 790, label = str_wrap("Recovery", 10) ,size = 2.5) +
  annotate("text", x = 48, y = 790, label = str_wrap("Lockdown", 10) ,size = 2.5) +
  geom_line(size = 1.1) +
  facet_wrap(~ID) +
  scale_y_continuous(limits = c(0, NA_real_), oob = squish) +
  scale_colour_manual(values = c("#f9bf07", "#5881c1", "#ec6555")) +
  theme(strip.text = element_text(face = "bold"),
        legend.position = c(0.8,0.2)) +
  labs(title = "Comparing regional trends to a London-style recovery highlights potentially unnecessary lost admissions",
       subtitle = "Elective hip admissions compared to a London-style modelled recovery: Admissions by NHSE region",
       x = "Week number", 
       y = "Admissions",
       colour = "",
       caption = "Note: Modelled trend matches the recovery profile of London NHSE region"
       )

```
<br>
Note: the modelled line is a London-style recovery, not Londons exact 2020 trend. The model had to return to the regions pre-pandemic level by the same date that London did, so London's curve could not simply be applied to the other regions as pre-pandemic rates differed greatly.

The area between the modelled trend and the 2020 trend represent the admissions that did not happen but could have if other regions recovery had matched that of London. 

A total of **19,013** admissions for elective hip replacements could have occurred between mid-March and the end of 2020, if other regions had achieved the same recovery profile as London.

<br>
<br>
```{r London-style recovery table, warning = FALSE, message = FALSE}
create_dt(
LTR_union_diff_london %>%
  summarise(Deficit = round(sum(diff_london), 0)) %>% 
  mutate(total = sum(Deficit)) %>% 
  rename(Region = id) %>% 
  select(-total) %>% 
  arrange(desc(Deficit))
)
```



# Notes and references

Notes: 

COVID-19 lockdown dates: <https://www.instituteforgovernment.org.uk/sites/default/files/timeline-lockdown-web.pdf>

References:

[1] NHS England., 2021. Consultant-led Referral to Treatment Waiting Times Data 2020-21. Available at: <https://www.england.nhs.uk/statistics/statistical-work-areas/rtt-waiting-times/rtt-data-2020-21/#Jan21> (Accessed 16th March 2021)

[2] Nikolova, S., Harrison, M. and Sutton, M., 2016. The impact of waiting time on health gains from surgery: Evidence from a national patientreported outcome dataset. Health economics, 25(8), pp.955-968. 



